---
title: "0_Processing_updated"
author: "Namita Joshi (using Lolitas script with updated data)"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    code_folding: show
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,message = FALSE,warning = FALSE)
```


# Load packages
```{r,echo=FALSE,message=FALSE,warning=FALSE}
library(stringr)
library(tidyr)
library(dplyr)
library(readr)
library(readxl)
library(ggplot2)
library(RColorBrewer)
library(patchwork)
library(DT)
library(ellmer)
```

# Preparation
To run this code, download the file "AdaptationElements" available at: https://doi.org/10.7910/DVN/VK3CP9.
Adjust the file path below with the correct location where you would saved the "AdaptationElements" file
```{r}

xlsx_path <- "data/BTRsMergedData _18Aug2025.xlsx"

adaptdata <- read_excel(xlsx_path, sheet = "DetailedInfo")
overview  <- read_excel(xlsx_path, sheet = "Overview")

```

# Step 1: Hazard type 

This section classifies climate-related hazards mentioned in the `ElementText` column of the `adaptdata` dataset using regex-based keyword matching, aligned with the protocol’s definitions.


1. **Column Creation**  
   A new column `HazardType` is added to store the hazard category for each entry. This column is positioned after `ElementText`.

2. **Keyword Mapping**  
   Hazard categories are mapped to regex patterns that represent keywords found in the protocol definitions. These keywords are matched case-insensitively within the `ElementText`.

3. **Tagging**  
   The script loops over all hazard patterns and tags any matching rows accordingly.

---

### Hazard Categories and Associated Keywords

```{r,echo=FALSE}
library(knitr)
library(kableExtra)

hazard_keywords <- tribble(
  ~"Hazard Category", ~"Keywords Used",
  
  "Extreme temperature", "heat wave, heatwave, excessive heat, high temperature, extreme temperature, extreme heat, extremely low, cold waves, cold wave, snow, ice, frost, freeze, severe winter, maximum and minimum, warmer, higher occurrence of hot, glacier, snowfall, evapo, extreme weather, heat stress, hot days, hot nights",
  
  "Storm", "storm, tropical storm, cyclone, cyclones, Cyclonic activity, typhoon, typhoons, hail, lightning, thunderstorm, heavy rain, windstorm, sand storm, dust storm, tornado, violent rains, torrential, strong winds",
  
  "Drought", "drought, drought cycle, prolonged droughts, drougths, dry spell, dry days, aridity",
  
  "Wildfire", "wildfire, wild fire, forest fire, land fire, bush fire, pasture fire, fire",
  
  "Landslide", "landslide, land slide",
  
  "Flood", "flood, coastal flood, riverine flood, flash flood, ice jam flood, inundation",
  
  "Change in temperature", "change in temperature, alteration in average temperature, temperature change, temperature rise, temperature drop, consistent change in temperature, rise in temperature, increase in temperature, increases in temperature, temperature increase, increasing temperature, increased temperature, increase temperatures, higher temperatures, average temperature, annual temperature, annual mean temperature, annual air temperature, minimum temperatures, maximum temperatures, number of warm days, rising temperatures, warming temperatures",
  
  "Change in precipitation", "change in precipitation, alteration in precipitation, precipitation patterns, shift in timing, shift in amount, shift in intensity, shift in frequency, rainfall, precipitation, distribution of rains, disruption of rains, fluctuation of rains, shorter rain, earlier and ending later",
  
  "Salinization", "salinization, salt content, saltwater, salt water, salinity, increase in salt content",
  
  "Land degradation", "land degradation, decline in land quality, decline in land health, pasture degradation, desertification, loss of organic matter, degradation of land, erosion, soil erosion",
  
  "Sea level rise", "sea level rise, sea level, increase in sea level, coastal flooding, coastal erosion, beach loss, coastline retreat, submersion, water mass",
  
  "Sea temperature", "sea temperature, sea surface temperature, change in sea temperature, ocean temperature, water temperature, surface temperature, seawater surface",
  
  "Ocean acidification", "ocean acidification, reduction in pH, acidity, acidification, acidic, coral",
  
  "Pest and disease", "pest, disease, epidemic, infestation, invasion, insect infestation, vector-borne, biological event, invasive species"
)

kable(hazard_keywords, "html", escape = FALSE) %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed", "responsive")) %>%
  column_spec(2, width = "50em")

```

```{r,echo=FALSE}
# 1) Initialize HazardType and relocate
adaptdata$HazardType <- ""
adaptdata <- adaptdata %>% relocate(HazardType, .after = ElementText)

# 2) Define your hazard → regex map
haz_patterns <- list(
  "Extreme temperature" = paste(
    "heat wave", "heatwave", "excessive heat", "high temperature", "extreme temperature",
    "extreme heat", "extremely low", "cold waves", "cold wave", "snow", "ice", "frost",
    "freeze", "severe winter", "maximum and minimum", "warmer", "higher occurrence of hot",
    "glacier", "snowfall", "evapo","extreme weather","heat stress","hot days","hot nights", sep = "|"
  ),
  
  "Storm" = paste(
    "storm", "tropical storm", "cyclone", "cyclones","Cyclonic activity","typhoon", "typhoons", "hail",
    "lightning", "thunderstorm", "heavy rain", "windstorm", "sand storm", "dust storm",
    "tornado", "violent rains","torrential","strong winds", sep = "|"
  ),
  
  "Drought" = paste(
    "drought", "drought cycle", "prolonged droughts","drougths","dry spell","dry days","aridity", sep = "|"
  ),
  
  "Wildfire" = paste(
    "wildfire", "wild fire", "forest fire", "land fire", "bush fire", "pasture fire","fire",
    sep = "|"
  ),
  
  "Landslide" = paste(
    "landslide", "land slide", sep = "|"
  ),
  
  "Flood" = paste(
    "\\bflood(s|ing)?\\b", "coastal flood", "riverine flood", "flash flood", "ice jam flood",
    "inundation", sep = "|"
  ),
  
  "Change in temperature" = paste(
    "change in temperature", "alteration in average temperature", "temperature change",
    "temperature rise", "temperature drop", "consistent change in temperature",
    "rise in temperature", "increase in temperature", "increases in temperature",
    "temperature increase", "increasing temperature", "increased temperature",
    "increase temperatures", "higher temperatures", "average temperature",
    "annual temperature","annual mean temperature","annual air temperature","minimum temperatures","maximum temperatures","number of warm days","rising temperatures","warming temperatures", sep = "|"
  ),
  
  "Change in precipitation" = paste(
    "change in precipitation", "alteration in precipitation", "precipitation patterns",
    "shift in timing", "shift in amount", "shift in intensity", "shift in frequency",
    "rainfall", "precipitation", "distribution of rains", "disruption of rains",
    "fluctuation of rains", "shorter rain", "earlier and ending later", sep = "|"
  ),
  
  "Salinization" = paste(
    "salinization", "salt content", "saltwater", "salt water", "salinity",
    "increase in salt content", sep = "|"
  ),
  
  "Land degradation" = paste(
    "land degradation", "decline in land quality", "decline in land health",
    "pasture degradation", "desertification", "loss of organic matter",
    "degradation of land", "erosion", "soil erosion", sep = "|"
  ),
  
  "Sea level rise" = paste(
    "sea[ -]?level rise", "sea level", "increase in sea level", "coastal flooding",
    "coastal erosion", "beach loss", "coastline retreat", "submersion","water mass","sea level rise", sep = "|"
  ),
  
  "Sea temperature" = paste(
    "sea temperature", "sea surface temperature", "change in sea temperature",
    "ocean temperature", "water temperature","surface temperature","seawater surface", sep = "|"
  ),
  
  "Ocean acidification" = paste(
    "ocean acidification", "reduction in pH", "acidity", "acidification", "acidic",
    "coral", sep = "|"
  ),
  
  "Pest and disease" = paste(
    "pest", "disease", "epidemic", "infestation", "invasion", "insect infestation",
    "vector-borne", "biological event,invasive species", sep = "|"
  )
)


# 3) Safe text (no NAs)
safe_text <- ifelse(is.na(adaptdata$ElementText), "", adaptdata$ElementText)

# 4) Minimal loop to append multiple types
for(type in names(haz_patterns)) {
  # find rows that are Hazard and match the pattern
  hits <- grepl(haz_patterns[[type]], safe_text, ignore.case = TRUE)
  idx  <- which(!is.na(adaptdata$Element) & adaptdata$Element == "Hazard" & hits)
  
  if (length(idx) > 0) {
    current <- adaptdata$HazardType[idx]
    adaptdata$HazardType[idx] <- ifelse(
      current == "",
      type,
      paste(current, type, sep = ";")
    )
  }
}

# Now any Hazard row matching multiple patterns will contain e.g. "Flood;Storm;Drought"

```

## Coverage of hazard type 

```{r}
# Count number of 'Hazard' elements that received a HazardType
assigned_hazards <- adaptdata %>%
  filter(Element == "Hazard" & HazardType != "") %>%
  nrow()

# Total number of Hazard elements
total_hazards <- adaptdata %>%
  filter(Element == "Hazard") %>%
  nrow()

# Display the result
cat("Hazard rows assigned a category:", assigned_hazards, "out of", total_hazards, "\n")
cat("**Coverage Rate:**", round(assigned_hazards / total_hazards * 100, 1), "%\n")
```

## Hazard type summary data 

```{r,echo=FALSE}
library(dplyr)
library(tidyr)
library(DT)

# 1. Filter for hazard elements with at least one category
hazard_counts <- adaptdata %>%
  filter(Element == "Hazard" & HazardType != "") %>%
  separate_rows(HazardType, sep = ";") %>%
  count(HazardType, sort = TRUE) %>%
  rename(`Hazard Category` = HazardType, `Occurrences` = n)

# 2. Display with DT for interactivity
DT::datatable(
  hazard_counts,
  rownames = FALSE,
  options = list(
    pageLength = 10,
    autoWidth = TRUE,
    dom = 'tip'
  )
)

write.csv(hazard_counts,   "hazard_counts.csv", row.names = FALSE)

#3. Visualize hazard count


hazard_plot <- ggplot(hazard_counts,
       aes(x = reorder(`Hazard Category`, Occurrences),
           y = Occurrences,
           fill = Occurrences)) +
  geom_col() +
  coord_flip() +
  scale_fill_viridis_c(option = "plasma", direction = -1, guide = "none") +
  labs(title = "Hazard categories by count", x = NULL, y = "Occurrences") +
  theme_minimal(base_size = 12) +
  theme(
    legend.position = "none",
    axis.text.y = element_text(size = 14),
    axis.text.x = element_text(size = 14)
  )

hazard_plot



```

## Hazard type raw data 

```{r}
# Load DT
library(DT)

# Filter Hazard elements and show them
hazard_table <- adaptdata %>%
  filter(Element == "Hazard") %>%
  select(Country, Document, ElementText, HazardType) %>%
  mutate(
    ElementText = ifelse(
      nchar(ElementText) > 50,
      paste0("<span title='", ElementText, "'>", substr(ElementText, 1, 50), "...</span>"),
      ElementText
    )
  )

# Display the table
datatable(
  hazard_table,
  escape = FALSE,  # Allow HTML rendering
  options = list(pageLength = 10, autoWidth = TRUE),
  caption = "📄 Hazard elements and their assigned HazardType"
)

write.csv(hazard_table,   "hazard_table.csv", row.names = FALSE)

```


# Step 2: Systems at risk

We assign a SystemType to each row of adaptdata where the element is "System at risk". This classification is based on the official reporting protocol definitions, which describe different climate-sensitive systems affected by hazards (e.g., agriculture, biodiversity, infrastructure).

To do this, we use a dictionary of keyword patterns that match text found in the ElementText column. These keywords are derived from the protocol and expanded where needed to capture variations in language.

Only rows where the element is "System at risk" are considered.

The following table summarizes the system categories used for classification and the corresponding keyword patterns:

```{r,echo=FALSE}
library(knitr)
library(kableExtra)
library(tibble)

system_table <- tribble(
  ~`System Type`, ~`Keywords Used`,

  "Crop", "crop, cropping systems, crop production, yield, cultivated area, agriculture, agricultural, agricultural production, agricultural pests, pest, irrigated, rain-fed, farming, land use planning, crop loss, agroecological zone, production",

  "Livestock", "livestock, pasture, pastoralist, pastoral area, grazing, animal health, productivity losses, herder, livestock loss",

  "Fisheries and aquaculture", "fish, fisheries, aquaculture, fishing, marine harvest",

  "Forest", "forest, forestry, forest product, tree, non-timber",

  "Terrestrial", "terrestrial ecosystem, terrestrial, drylands, land resource, natural resource, ecosystem structure, ecosystem services, ecological system, desertification, land degradation, soil degradation, soil erosion, environment",

  "Freshwater", "freshwater, wetlands, inland wetlands, water resource, drinking water, potable water, water quality, water availability, water supply, river, water scarcity, water shortage, hydrological cycle, water stress, water table, water source, groundwater, water supplied, dams",

  "Biodiversity", "biodiversity, flora, fauna, species, species extinction, extinction, range of species, ecosystem change",

  "Coastal", "coast, coastal, marine, mangrove, ocean, coral, beach, coastal erosion, sea level rise, blue carbon, coastal ecosystem, coastal zone, eutrophication, algal bloom",

  "Food and nutrition", "food security, food insecure, food insecurity, nutrition, malnutrition, hunger, food safety, food availability, famine, undernutrition, overnutrition, obesity",

  "Gender and inclusion", "gender, women, youth, children, elderly, inclusion, social exclusion, vulnerable group, vulnerable population, minority group, indigenous, small-scale producer, pastoralist, fishing communities, forest-based communities, high-risk regions",

  "Livelihoods and poverty", "livelihood, poverty, income, employment, labor, economic activity, loss of income, loss of livelihood, safety net, insurance, socio-economic development, tourism, subsistence, workforce, economic, tourists, outdoor activities, ski, vacation",

  "Health", "health, mental health, morbidity, mortality, vector-borne disease, water-borne disease, infectious disease, respiratory disease, malaria, epidemic, climate-sensitive disease, heat morbidity, disease, deaths, human lives, pollution, life expectancy",

  "Infrastructure and services", "infrastructure, critical infrastructure, services, critical services, road, bridge, electricity, power supply, energy, water supply, sanitation, hygiene, education, school, building, housing, settlement, evacuation, telecom, transport, waste management, power, hydropower, railways, port, industry, material, industries",

  "Human security and Peace", "migration, displacement, conflict, armed conflict, national security, human security, organized conflict, climate-induced migration, refugee, peace"
)

kable(system_table, "html", escape = FALSE, align = "l") %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed", "responsive"), full_width = FALSE) %>%
  column_spec(2, width = "50em", extra_css = "word-wrap: break-word; white-space: normal;")

```

```{r,echo=FALSE}
# 1) Initialize SystemType and relocate
adaptdata$SystemType <- ""
adaptdata <- adaptdata %>% relocate(SystemType, .after = HazardType)

# 2) Define your system‐to‐pattern map
sys_patterns <- list(
  Crop = paste(
    "crop", "cropping systems", "crop production", "yield", "cultivated area",
    "agriculture", "agricultural", "agricultural production", "agricultural pests",
    "pest", "irrigated", "rain-fed", "farming", "land use planning", "crop loss","agroecological zone","production", sep = "|"
  ),
  
  Livestock = paste(
    "livestock", "pasture", "pastoralist", "pastoral area", "grazing", 
    "animal health", "productivity losses", "herder", "livestock loss", sep = "|"
  ),
  
  `Fisheries and aquaculture` = paste(
    "fish", "fisheries", "aquaculture", "fishing", "marine harvest", sep = "|"
  ),
  
  Forest = paste(
    "forest", "forestry", "forest product", "tree", "non-timber", sep = "|"
  ),
  
  Terrestrial = paste(
    "terrestrial ecosystem", "terrestrial", "drylands", "land resource", 
    "natural resource", "ecosystem structure", "ecosystem services",
    "ecological system", "desertification", "land degradation", 
    "soil degradation", "soil erosion", "environment", sep = "|"
  ),
  
  Freshwater = paste(
    "freshwater", "wetlands", "inland wetlands", "water resource", "drinking water",
    "potable water", "water quality", "water availability", "water supply",
    "river", "water scarcity", "water shortage", "hydrological cycle",
    "water stress", "water table", "water source", "groundwater","water supplied","dams", sep = "|"
  ),
  
  Biodiversity = paste(
    "biodiversity", "flora", "fauna", "species", "species extinction",
    "extinction", "range of species", "ecosystem change", sep = "|"
  ),
  
  Coastal = paste(
    "coast", "coastal", "marine", "mangrove", "ocean", "coral", 
    "beach", "coastal erosion", "sea level rise", "blue carbon", 
    "coastal ecosystem", "coastal zone","eutrophication","algal bloom", sep = "|"
  ),
  
  `Food and nutrition` = paste(
    "food security", "food insecure", "food insecurity", "nutrition", 
    "malnutrition", "hunger", "food safety", "food availability", 
    "famine", "undernutrition", "overnutrition", "obesity", sep = "|"
  ),
  
  `Gender and inclusion` = paste(
    "gender", "women", "youth", "children", "elderly", "inclusion",
    "social exclusion", "vulnerable group", "vulnerable population",
    "minority group", "indigenous", "small-scale producer", 
    "pastoralist", "fishing communities", "forest-based communities",
    "high-risk regions", sep = "|"
  ),
  
  `Livelihoods and poverty` = paste(
    "livelihood", "poverty", "income", "employment","labor", "economic activity",
    "loss of income", "loss of livelihood", "safety net","insurance", 
    "socio-economic development", "tourism", "subsistence", "workforce","economic","tourism","tourists","outdoor activities","ski","vacation",
    sep = "|"
  ),
  
  Health = paste(
    "health", "mental health", "morbidity", "mortality", "vector-borne disease",
    "water-borne disease", "infectious disease", "respiratory disease", 
    "malaria", "epidemic", "climate-sensitive disease", "heat morbidity","disease","deaths","human lives","pollution","life expectancy", sep = "|"
  ),
  
  `Infrastructure and services` = paste(
    "infrastructure", "critical infrastructure", "services", "critical services",
    "road", "bridge", "electricity", "power supply", "energy", 
    "water supply", "sanitation", "hygiene", "education", "school", 
    "building", "housing", "settlement", "evacuation", "telecom", 
    "transport", "waste management","power","hydropower","railways","port","industry","material","industries", sep = "|"
  ),
  
  `Human security and Peace` = paste(
    "migration", "displacement", "conflict", "armed conflict", 
    "national security", "human security", "organized conflict", 
    "climate-induced migration", "refugee", "peace", sep = "|"
  )
)


# 3) Build a safe text vector (no NAs)
safe_text <- ifelse(is.na(adaptdata$ElementText), "", adaptdata$ElementText)

# 4) Loop through patterns, appending multiple matches with ";"
for(type in names(sys_patterns)) {
  # identify rows to tag
  idx <- which(
    adaptdata$Element == "System at risk" &
    grepl(sys_patterns[[type]], safe_text, ignore.case = TRUE)
  )
  
  if(length(idx)) {
    current <- adaptdata$SystemType[idx]
    adaptdata$SystemType[idx] <- ifelse(
      current == "",
      type,
      paste(current, type, sep = ";")
    )
  }
}

# Now any "System at risk" row matching multiple patterns will have
# e.g. "Crop;Livestock;Food and nutrition" in SystemType.

```

## Coverage of system at risk type 

```{r}
# Compute number of System at risk rows
total_system_rows <- sum(adaptdata$Element == "System at risk", na.rm = TRUE)

# Count how many received a classification
assigned_system_rows <- sum(adaptdata$Element == "System at risk" & adaptdata$SystemType != "", na.rm = TRUE)

# Print basic coverage stats
cat("**System at Risk Coverage:**", assigned_system_rows, "of", total_system_rows, 
    "rows were successfully assigned a category.\n")
cat("**Coverage Rate:**", round(assigned_system_rows / total_system_rows * 100, 1), "%\n")

```
## System at risk type summary data 

```{r,echo=FALSE}
library(dplyr)
library(tidyr)
library(DT)

# 1. Filter and separate multiple SystemType values
system_counts <- adaptdata %>%
  filter(Element == "System at risk" & SystemType != "") %>%
  separate_rows(SystemType, sep = ";") %>%
  count(SystemType, sort = TRUE) %>%
  rename(`System at Risk Category` = SystemType, `Occurrences` = n)

# 2. Display the interactive table
DT::datatable(
  system_counts,
  rownames = FALSE,
  options = list(
    pageLength = 10,
    autoWidth = TRUE,
    dom = 'tip'
  )
)

write.csv(system_counts,   "system_counts.csv", row.names = FALSE)

# 3. Visualize systems at risk
system_plot <- ggplot(system_counts,
       aes(x = reorder(`System at Risk Category`, Occurrences),
           y = Occurrences,
           fill = `System at Risk Category`)) +
  geom_col(show.legend = FALSE) +
  coord_flip() +
  scale_fill_viridis_d(option = "mako") +
  labs(title = "Systems at risk by count", x = NULL, y = "Occurrences") +
  theme_minimal(base_size = 12) +
  theme(
    axis.text.y = element_text(size = 14),
    axis.text.x = element_text(size = 14)
  )

system_plot


```

## System at risk type raw data 

```{r,echo=FALSE}
# Install if needed: install.packages("DT")
library(DT)


# Show only tagged 'System at risk' rows and format long text for tooltip
tagged_systems <- adaptdata %>%
  filter(Element == "System at risk", SystemType != "") %>%
  mutate(
    ElementText = ifelse(
      nchar(ElementText) > 50,
      paste0("<span title='", ElementText, "'>", substr(ElementText, 1, 50), "...</span>"),
      ElementText
    )
  )

# Display the table with tooltips
DT::datatable(
  tagged_systems,
  escape = FALSE,  # Allow HTML
  options = list(pageLength = 10, autoWidth = TRUE),
  caption = "📌 Tagged 'System at risk' Rows with Assigned Categories"
)



```

# Step 3: Sectors
We classify sectors in two ways:

1. **SectorType (IPCC categories)** — based on keywords found in the `Sector` and `SystemType` columns. This follows the sector taxonomy defined in the MPGs (Table 1).
2. **SectorType_GGA (GGA themes)** — which maps the sectors into broader Global Goal on Adaptation (GGA) thematic groups as per the protocol.

The tagging logic first applies regular expressions to the text in `Sector`, with some fallback rules based on the `SystemType` when relevant. The classification is additive and accounts for overlapping sector concepts.

Below is a summary table of sector categories and the associated keywords used.


### SectorType (IPCC categories) - Keywords used

```{r,echo=FALSE}


sector_table <- tribble(
  ~`IPCC Sector Category`, ~`Keywords / Match Terms`,

  "Food, fiber and other ecosystem products",
  "agri, agro, agriculture, food, crop, livestock, animal, fish, fisheries, aquacultur, aquaculture, seed, irrigation, value chain, land use, land tenure, land and forestry, agroforest, Agriculture and food security, Agriculture, Climate services, Others, forest and other land uses, Agriculture, forest and other land uses, land affairs, land reforms, Food and nutrition, Crop, Livestock, Fisheries and aquaculture, Sustainable development, Agriculture",

  "Terrestrial and freshwater ecosystems",
  "forest, environ, environment, enviornment, ecosystem, biodiversity, natural, ecology, ecolog, wildlife, REDD, peatland, protected area, Terrestrial, Freshwater, land",

  "Ocean and coastal ecosystems",
  "ocean, marine, coast, coastal, coastal land use, coastal zone, blue carbon, mangrove, Tourism and Coastal Zone Management, Coastal zone management",

  "Water, sanitation and hygiene",
  "sanitation, water and sanitation, sewerage, hygiene, water use, water security, water and energy, water, sanitation and waste",

  "Cities, settlements and key infrastructure",
  "city, cities, urban, settlement, infrastructure, housing, habitat, industr, waste, transport, energy, landfills, sanitary landfills, mining, mineral resources, mineral products, telecommunications, Infrastructure, transport and building, Infrastructure and services, Cities and Built Environment, Land Use and Human Settlements Development, Habitat, urban planning and development of the territory, Housing, Territorial Development and Urban Planning, Renewable Energy, Urban planning and infrastructure, Urban Development & Tourism, Private sector/trade; Manufacturing; Business process",

  "Health, wellbeing and communities",
  "health, well-being, wellbeing, nutrition, culture, territorial communit, territorial communities, local knowledge, Human security, Vulnerable communities, Territorial development",

  "Livelihoods, poverty and sustainable development",
  "social, poverty, people, econom, capacity, education, employment, tourism, rural development, sustainable development, economic and social infrastructure, Livelihoods and poverty, protection, social protection, transfer, income, revenue, social infrastructures, Education, research, Education, training, research, Multiple: Social Economy, Tourism, Multiple: Social Affairs, Women and Family, Multiple: Planning, Rural development, Multiple: Sustainable development, Planning",

  "Crosscutting",
  "cross-cutting, cross cutting, cross-sectoral, innovation, research, R&D, integration, empower, gender, women, youth, disaster, risk, climate, climate service, meteo, warning, governance, legislation, policy, policies, institution, M&E, devolution, private sector, public sector, territory, spatial planning, planning, weather",

  "Not specified",
  "Not specified"
)

kable(sector_table, "html", escape = FALSE, align = "l") %>%
  kable_styling(full_width = FALSE,
                bootstrap_options = c("striped", "hover", "condensed", "responsive")) %>%
  column_spec(2, width = "50em",
              extra_css = "word-wrap: break-word; white-space: normal;")

```


```{r,echo=FALSE}


# Patterns per protocol + your extra phrases/stems
ipcc_patterns <- list(
  "Food, fiber and other ecosystem products" =
    paste0(
      "(",
      paste(c(
        # protocol terms
        "agri","agro","agriculture","food","crop","livestock","animal","fish",
        "fisheries","aquacultur","aquaculture","seed","irrigation","value chain",
        "land use","land tenure","land and forestry","agroforest",
        # extras seen in your prior rules
        "Agriculture,? (?:forest|and other land uses)","land affairs","land reforms",
        "Food and nutrition","Crop","Livestock","Fisheries and aquaculture",
        "Agriculture and food security","Agriculture, Climate services, Others",
        "Sustainable development, Agriculture"
      ), collapse="|"),
      ")"
    ),

  "Terrestrial and freshwater ecosystems" =
    paste0(
      "(",
      paste(c(
        # protocol terms
        "forest","environ","environment","ecosystem","biodiversity","natural","ecolog",
        "wildlife","REDD","peatland","protected area",
        # extras (including misspelling & broad terms you used)
        "enviornment","Terrestrial","Freshwater","Biodiversity","land\\b"
      ), collapse="|"),
      ")"
    ),

  "Ocean and coastal ecosystems" =
    paste0(
      "(",
      paste(c(
        # protocol terms
        "ocean","marine","coast","coastal","coastal land use","coastal zone","blue carbon","mangrove",
        # extras
        "Tourism and Coastal Zone Management","Coastal zone management"
      ), collapse="|"),
      ")"
    ),

  "Water, sanitation and hygiene" =
    paste0(
      "(",
      paste(c(
        # protocol terms
        "sanitation","water and sanitation","sewerage","hygiene",
        # extras
        "water security","water use","water and energy","water, sanitation and waste"
      ), collapse="|"),
      ")"
    ),

  "Cities, settlements and key infrastructure" =
    paste0(
      "(",
      paste(c(
        # protocol terms
        "city","cities","urban","settlement","infrastructure","housing","habitat",
        "industr","waste","transport","energy",
        # extras (phrases & sectors you mapped before)
        "Infrastructure, transport and building","Infrastructure and services",
        "landfills","sanitary landfills","mining","mineral resources","mineral products",
        "telecommunications","Telecommunications",
        "Cities and Built Environment","Land Use and Human Settlements Development",
        "Habitat, urban planning and development of the territory",
        "Housing, Territorial Development and Urban Planning",
        "Renewable Energy","Urban planning and infrastructure",
        "Urban Development & Tourism","Private sector/trade; Manufacturing; Business process"
      ), collapse="|"),
      ")"
    ),

  "Health, wellbeing and communities" =
    paste0(
      "(",
      paste(c(
        # protocol terms
        "health","well-?being","wellbeing","nutrition","culture",
        "territorial communit","local knowledge",
        # extras you used/mapped
        "territorial communities","Human security","Vulnerable communities","Territorial development"
      ), collapse="|"),
      ")"
    ),

  "Livelihoods, poverty and sustainable development" =
    paste0(
      "(",
      paste(c(
        # protocol terms
        "social","poverty","people","econom","capacity","education","employment",
        "tourism","rural development","sustainable development","economic and social infrastructure",
        # extras
        "Livelihoods and poverty","protection","social protection","transfer",
        "income","revenue","social infrastructures",
        "Education, research","Education, training, research",
        "Multiple: Social Economy, Tourism","Multiple: Social Affairs, Women and Family",
        "Multiple: Planning, Rural development","Multiple: Sustainable development, Planning"
      ), collapse="|"),
      ")"
    ),

  "Crosscutting" =
    paste0(
      "(",
      paste(c(
        # protocol terms
        "cross[- ]?cutting","cross[- ]?sectoral","innovation","research","R&D","integration",
        "empower","gender","women","youth","disaster","risk","climate( service)?",
        "meteo","warning","governance","legislation","polic(?:y|ies)","institution","M&E",
        # extras
        "devolution","private sector","public sector","territory","spatial planning","\\bplanning\\b"
      ), collapse="|"),
      ")"
    )
)

# Tie-break priority = list order above (adjust if you want a different precedence)
ipcc_priority <- names(ipcc_patterns)

# Ensure the column exists and sits after Sector
if (!"SectorType" %in% names(adaptdata)) adaptdata$SectorType <- NA_character_
adaptdata <- adaptdata %>% relocate(SectorType, .after = Sector)

# Build the text to scan (Sector first, with SystemType as context/fallback)
adaptdata <- adaptdata %>%
  mutate(
    .ipcc_text = paste0(
      coalesce(as.character(Sector), ""), " ",
      coalesce(as.character(SystemType), "")
    )
  ) %>%
  rowwise() %>%
  mutate(
    # count matches per sector
    .counts = list(vapply(ipcc_patterns, function(pat) {
      str_count(.ipcc_text, regex(pat, ignore_case = TRUE))
    }, integer(1))),
    # pick a single best sector; NA if no matches
    SectorType = {
      counts <- unlist(.counts)
      if (all(counts == 0L)) NA_character_ else {
        winners <- which(counts == max(counts))
        ipcc_priority[winners[1]]
      }
    }
  ) %>%
  ungroup() %>%
  select(-.ipcc_text, -.counts)

```


## Sector GGA

```{r, echo=FALSE}


gga_table <- tribble(
  ~`GGA Sector Theme`, ~`Keywords / Match Terms`,

  "Water and sanitation",
  "sanitation, water and sanitation, sewerage, hygiene, water security, water use, water and energy, water, sanitation and waste",

  "Food and agriculture",
  "agri, agro, agriculture, food, crop, livestock, animal, fish, fisheries, aquacultur, aquaculture, seed, irrigation, value chain, land use, land tenure, land and forestry, agroforest, nutrition, food secur, Agriculture and food security, Food and nutrition, Agriculture, Climate services, Others, Sustainable development, Agriculture, land affairs, land reforms, Agriculture, forest and other land uses",

  "Health",
  "health, well-being, wellbeing, territorial communit, territorial communities, local knowledge, Human security, Vulnerable communities, Territorial development",

  "Biodiversity and ecosystems",
  "ocean, marine, coast, coastal land use, coastal zone, blue carbon, mangrove, forest, environ, environment, enviornment, ecosystem, biodiversity, natural, ecology, ecolog, wildlife, REDD, peatland, protected area, freshwater, Terrestrial, Tourism and Coastal Zone Management, Coastal zone management",

  "Infrastructure and human settlements",
  "city, cities, urban, settlement, infrastructure, housing, habitat, industr, waste, transport, energy, landfills, sanitary landfills, mining, mineral resources, mineral products, telecommunications, Infrastructure, transport and building, Infrastructure and services, Cities and Built Environment, Land Use and Human Settlements Development, Habitat, urban planning and development of the territory, Housing, Territorial Development and Urban Planning, Renewable Energy, Urban planning and infrastructure, Urban Development & Tourism, spatial planning, planning, territory, Private sector/trade; Manufacturing; Business process, private sector, public sector",

  "Poverty eradication and livelihoods",
  "social, poverty, people, econom, capacity, education, employment, tourism, rural development, sustainable development, economic and social infrastructure, livelihood, income, revenue, social protection, protection, transfer, social infrastructures, Education, research, Education, training, research, Multiple: Social Economy, Tourism, Multiple: Social Affairs, Women and Family, Multiple: Planning, Rural development, Multiple: Sustainable development, Planning",

  "Cultural heritage",
  "cultur, cultural heritage, heritage site, traditional knowledge, local knowledge, indigenous"
)

kable(gga_table, "html", escape = FALSE, align = "l") %>%
  kable_styling(full_width = FALSE,
                bootstrap_options = c("striped", "hover", "condensed", "responsive")) %>%
  column_spec(2, width = "50em",
              extra_css = "word-wrap: break-word; white-space: normal;")


```


```{r,echo=FALSE}



# Build regex patterns from the table above (case-insensitive)
gga_patterns <- list(
  "Water and sanitation" =
    paste0("(", paste(c(
      "sanitation","water and sanitation","sewerage","hygiene",
      "water security","water use","water and energy","water, sanitation and waste"
    ), collapse="|"), ")"),

  "Food and agriculture" =
    paste0("(", paste(c(
      "agri","agro","agriculture","food","crop","livestock","animal","fish","fisheries",
      "aquacultur","aquaculture","seed","irrigation","value chain","land use","land tenure",
      "land and forestry","agroforest","nutrition","food secur",
      "Agriculture and food security","Food and nutrition",
      "Agriculture, Climate services, Others","Sustainable development, Agriculture",
      "land affairs","land reforms","Agriculture, forest and other land uses"
    ), collapse="|"), ")"),

  "Health" =
    paste0("(", paste(c(
      "health","well-?being","wellbeing","territorial communit","territorial communities",
      "local knowledge","Human security","Vulnerable communities","Territorial development"
    ), collapse="|"), ")"),

  "Biodiversity and ecosystems" =
    paste0("(", paste(c(
      "ocean","marine","coast","coastal land use","coastal zone","blue carbon","mangrove",
      "forest","environ","environment","enviornment","ecosystem","biodiversity","natural",
      "ecology","ecolog","wildlife","REDD","peatland","protected area","freshwater",
      "Terrestrial","Tourism and Coastal Zone Management","Coastal zone management"
    ), collapse="|"), ")"),

  "Infrastructure and human settlements" =
    paste0("(", paste(c(
      "city","cities","urban","settlement","infrastructure","housing","habitat","industr",
      "waste","transport","energy","landfills","sanitary landfills","mining","mineral resources",
      "mineral products","telecommunications","Infrastructure, transport and building",
      "Infrastructure and services","Cities and Built Environment",
      "Land Use and Human Settlements Development",
      "Habitat, urban planning and development of the territory",
      "Housing, Territorial Development and Urban Planning",
      "Renewable Energy","Urban planning and infrastructure","Urban Development & Tourism",
      "spatial planning","\\bplanning\\b","territory",
      "Private sector/trade; Manufacturing; Business process","private sector","public sector"
    ), collapse="|"), ")"),

  "Poverty eradication and livelihoods" =
    paste0("(", paste(c(
      "social","poverty","people","econom","capacity","education","employment","tourism",
      "rural development","sustainable development","economic and social infrastructure",
      "livelihood","income","revenue","social protection","protection","transfer",
      "social infrastructures","Education, research","Education, training, research",
      "Multiple: Social Economy, Tourism","Multiple: Social Affairs, Women and Family",
      "Multiple: Planning, Rural development","Multiple: Sustainable development, Planning"
    ), collapse="|"), ")"),

  "Cultural heritage" =
    paste0("(", paste(c("cultural heritage","heritage site","traditional knowledge","local knowledge","indigenous"
    ), collapse="|"), ")")
)

# Priority order for tie-breaks (edit if you prefer a different precedence)
gga_priority <- names(gga_patterns)

# Ensure column exists
if (!"SectorType_GGA" %in% names(adaptdata)) adaptdata$SectorType_GGA <- NA_character_

# Build text to scan (Sector first, with SystemType as context)
adaptdata <- adaptdata %>%
  mutate(
    .gga_text = paste0(
      coalesce(as.character(Sector), ""), " ",
      coalesce(as.character(SystemType), "")
    )
  ) %>%
  rowwise() %>%
  mutate(
    .counts_gga = list(vapply(gga_patterns, function(pat) {
      str_count(.gga_text, regex(pat, ignore_case = TRUE))
    }, integer(1))),
    SectorType_GGA = {
      cnt <- unlist(.counts_gga)
      if (all(cnt == 0L)) NA_character_ else {
        winners <- which(cnt == max(cnt))
        gga_priority[winners[1]]
      }
    }
  ) %>%
  ungroup() %>%
  select(-.gga_text, -.counts_gga)


```

## Sector coverage 

```{r}
# Count rows classified under SectorType
total_sector_rows <- sum(!is.na(adaptdata$Sector))
classified_sector_rows <- sum(!is.na(adaptdata$SectorType))
classified_sector_rows_gga <- sum(!is.na(adaptdata$SectorType_GGA))

cat("**SectorType (IPCC):**", classified_sector_rows, "of", total_sector_rows, "rows tagged\n")
cat("**SectorType_GGA (GGA):**", classified_sector_rows_gga, "of", total_sector_rows, "rows tagged\n")
cat("**Coverage Rate:**", round(classified_sector_rows / total_sector_rows * 100, 1), "%\n\n")



```
## Sector type summary 

```{r,echo=FALSE}

# --- IPCC Sector Type Frequency ---
sector_ipcc_counts <- adaptdata %>%
  filter(!is.na(SectorType))%>%
  count(SectorType, sort = TRUE) %>%
  rename(`IPCC Sector Category` = SectorType, `Occurrences` = n)

# --- GGA Sector Type Frequency ---
sector_gga_counts <- adaptdata %>%
  filter(!is.na(SectorType_GGA)) %>%
  count(SectorType_GGA, sort = TRUE) %>%
  rename(`GGA Sector Category` = SectorType_GGA, `Occurrences` = n)

# --- Display both tables side by side (optional tweak: two columns, one below the other if rendering limits apply) ---

# IPCC Sector Table
DT::datatable(
  sector_ipcc_counts,
  rownames = FALSE,
  caption = "📊 IPCC Sector Categories (SectorType)",
  options = list(pageLength = 10, autoWidth = TRUE, dom = 'tip')
)


# GGA Sector Table
DT::datatable(
  sector_gga_counts,
  rownames = FALSE,
  caption = "📊 GGA Sector Themes (SectorType_GGA)",
  options = list(pageLength = 10, autoWidth = TRUE, dom = 'tip')
)
write.csv(sector_ipcc_counts,   "sector_ipcc_counts.csv", row.names = FALSE)
write.csv(sector_gga_counts,   "sector_gga_counts.csv", row.names = FALSE)

# visualize sector type 
# IPCC
p_ipcc <- ggplot(sector_ipcc_counts,
                 aes(x = reorder(`IPCC Sector Category`, Occurrences),
                     y = Occurrences, fill = Occurrences)) +
  geom_col() +
  coord_flip() +
  scale_fill_viridis_c(option = "mako", guide = "none") +   
  labs(title = "IPCC sectors (SectorType)", x = NULL, y = "Occurrences") +
  theme_minimal(base_size = 12)

# GGA
p_gga <- ggplot(sector_gga_counts,
                aes(x = reorder(`GGA Sector Category`, Occurrences),
                    y = Occurrences, fill = Occurrences)) +
  geom_col() +
  coord_flip() +
  scale_fill_viridis_c(option = "cividis", guide = "none") + 
  labs(title = "GGA sectors (SectorType_GGA)", x = NULL, y = "Occurrences") +
  theme_minimal(base_size = 12)

# Panel + bigger axis text
sector_plot <- (p_ipcc | p_gga) &
  theme(
    legend.position = "none",
    axis.text.x = element_text(size = 14),
    axis.text.y = element_text(size = 14)
  )

sector_plot
```

## Sector type raw data 

```{r,echo=FALSE}

# Prepare the data with hover tooltips for long 'Sector' values
sector_tagged <- adaptdata %>%
  filter(!is.na(SectorType) | !is.na(SectorType_GGA)) %>%   # CHANGED
  mutate(
    Sector = ifelse(
      !is.na(Sector) & nchar(Sector) > 50,
      paste0("<span title='", Sector, "'>", substr(Sector, 1, 50), "...</span>"),
      Sector
    )
  )

# Render the datatable with HTML tooltips
datatable(
  sector_tagged,
  escape = FALSE,  # Allow HTML for tooltips
  options = list(pageLength = 10, autoWidth = TRUE),
  caption = "🏷️ Tagged Sector Rows with IPCC and GGA Classification"
)



```

# Step 4 : Climate Impact

## 4.1 GPT Extraction — impact keywords

- Each *Climate impact* passage is reduced by GPT to **short exact phrases** of consequences  
  (e.g., *“water scarcity”; “crop yield loss”; “heat-related illness”*).  
- Hazards (*“storm”*) and actions (*“irrigation”*) are excluded unless phrased as impacts  
  (e.g., *“increased flooding”*).  


## 4.2 Categorization — from phrases to classes

- Phrases are mapped to a fixed set of **impact categories**  
    - Water availability shortfall  
    - Water quality deterioration  
    - Crop production loss  
    - Livestock impacts  
    - Fisheries & aquaculture decline  
    - Food security & nutrition decline  
    - Human health burden  
    - Mental health impacts  
    - Healthcare system strain  
    - Mortality & injury  
    - Displacement & migration  
    - Damage to buildings & infrastructure  
    - Energy system impacts  
    - Transport disruption  
    - Supply chain & logistics disruptions  
    - Economic & livelihood losses  
    - Tourism & recreation impacts  
    - Ecosystem degradation & biodiversity loss  
    - Soil erosion & land degradation  
    - Marine ecosystem stressors  
    - Cultural heritage impacts  
    - Service delivery disruptions  
    - Carbon storage & productivity decline  
    - Coastal inundation & shoreline erosion  
    - Cryosphere degradation & ground instability  
    - Air quality deterioration  
    - Agricultural pests & diseases  
    - Agricultural system shifts  
    - Potential gains


- Each category has sample keywords + regex rules to ensure consistent matches.  
- Rows with only hazards (no consequences) are dropped.  
- Multiple categories can apply; results stored in `ImpactCategories`.  


```{r,echo=FALSE}

# suppressPackageStartupMessages({
#   library(dplyr)
#   library(stringr)
#   library(glue)
#   library(DT)
#   library(ellmer)
#   library(tidyr)
# })
# 
# # =================== Config ===================
# MODEL             <- "gpt-5-nano"
# RUN_TAG           <- "gpt5nano"                     # keep the SAME tag used last time
# OUTPUT_DIR        <- "out"
# OUTPUT_CSV        <- file.path(OUTPUT_DIR, paste0("climate_impact_keywords_", RUN_TAG, ".csv"))
# ERROR_LOG         <- file.path(OUTPUT_DIR, paste0("ci_errors_", RUN_TAG, ".log"))
# SAVE_EVERY        <- 25
# MAX_CHARS         <- 6000
# RPM               <- 60
# 
# RESUME_FROM_PREV  <- TRUE                           # ✅ read previous CSV and prefill
# RESET_PREVIOUS    <- FALSE                          # ✅ do NOT wipe already-filled values
# 
# # ---- API key (ellmer reads OPENAI_API_KEY) ----
# btr_key <- Sys.getenv("BTR_KEY")
# stopifnot("BTR_KEY is not set in your .Renviron (or current env)." = nzchar(btr_key))
# Sys.setenv(OPENAI_API_KEY = btr_key)
# 
# # =================== Utilities ===================
# dir.create(OUTPUT_DIR, showWarnings = FALSE, recursive = TRUE)
# 
# safe_write_csv <- function(df, path) {
#   tmp <- paste0(path, ".tmp")
#   ok  <- tryCatch({
#     utils::write.csv(df, tmp, row.names = FALSE)
#     TRUE
#   }, error = function(e) {
#     message("⚠ write.csv to tmp failed: ", conditionMessage(e)); FALSE
#   })
#   if (!ok || !file.exists(tmp)) {
#     alt <- file.path(dirname(path), paste0("backup_", format(Sys.time(), "%Y%m%d_%H%M%S"), ".csv"))
#     utils::write.csv(df, alt, row.names = FALSE)
#     message("✳ Wrote backup to: ", alt)
#     return(invisible(FALSE))
#   }
#   if (file.exists(path)) suppressWarnings(try(unlink(path), silent = TRUE))
#   ok2 <- suppressWarnings(try(file.rename(tmp, path), silent = TRUE))
#   if (identical(ok2, FALSE)) { file.copy(tmp, path, overwrite = TRUE); unlink(tmp) }
#   TRUE
# }
# 
# log_error <- function(msg) {
#   try({
#     cat(sprintf("[%s] %s\n", format(Sys.time(), "%Y-%m-%d %H:%M:%S"), msg),
#         file = ERROR_LOG, append = TRUE)
#   }, silent = TRUE)
# }
# 
# make_preview_html <- function(txt) {
#   ifelse(nchar(txt) > 90,
#          paste0("<span title='", txt, "'>", substr(txt, 1, 90), "...</span>"),
#          txt)
# }
# 
# clip_text <- function(x, max_chars = MAX_CHARS) {
#   if (is.na(x) || !nzchar(x) || nchar(x) <= max_chars) return(x)
#   head <- substr(x, 1, round(max_chars * 0.7))
#   tail <- substr(x, nchar(x) - round(max_chars * 0.3) + 1, nchar(x))
#   paste0(head, "\n...\n", tail)
# }
# 
# make_ci_prompt <- function(txt) {
#   glue::glue(
# 'Read the "Climate impact" passage.
# 
# Extract short key keywords that describe the CONSEQUENCES/IMPACTS of climate (what happens to people, ecosystems, services, assets).
# Rules:
# - Use EXACT short phrases from the text (same language).
# - Examples: "water shortage"; "crop yield loss"; "infrastructure damage"; "heat-related illness"; "biodiversity loss"; "coastal erosion".
# - Exclude drivers/hazards (e.g., "heat wave", "drought") unless the text itself presents them as the impact (e.g., "increased flooding").
# - Exclude actions/measures (e.g., "irrigation expansion", "early warning system").
# - Return ONE line only: keywords separated by "; ".
# 
# TEXT:
# """{clip_text(txt)}"""'
#   )
# }
# 
# rate_wrap <- function(f) f
# if (is.finite(RPM) && requireNamespace("ratelimitr", quietly = TRUE)) {
#   rate_wrap <- function(f) ratelimitr::limit_rate(f, ratelimitr::rate(n = RPM, period = 60))
# }
# 
# chat_client <- ellmer::chat_openai(model = MODEL)
# extract_once <- function(txt) chat_client$chat(make_ci_prompt(txt))
# 
# extract_ci_keywords <- rate_wrap(function(txt) {
#   if (is.na(txt) || !nzchar(txt)) return(NA_character_)
#   tries <- 0; last_err <- ""
#   repeat {
#     tries <- tries + 1
#     out <- try(extract_once(txt), silent = TRUE)
#     if (!inherits(out, "try-error")) {
#       out <- trimws(out)
#       return(if (nzchar(out)) out else NA_character_)
#     }
#     last_err <- conditionMessage(attr(out, "condition"))
#     if (grepl("context|token|length|too long|maximum", last_err, ignore.case = TRUE)) {
#       txt <- clip_text(txt, max_chars = round(MAX_CHARS * 0.7))
#     }
#     if (tries >= 5) { log_error(paste("API error after retries:", last_err)); return(NA_character_) }
#     Sys.sleep(min(runif(1, 1, 2) * (2^(tries - 1)), 30))
#   }
# })
# 
# # =================== Prepare data ===================
# if (!"RowID" %in% names(adaptdata)) adaptdata$RowID <- seq_len(nrow(adaptdata))
# if (!"ClimateImpact_Keywords" %in% names(adaptdata)) adaptdata$ClimateImpact_Keywords <- NA_character_
# 
# ci_idx_all <- which(grepl("^\\s*Climate impact\\s*$", adaptdata$Element, ignore.case = TRUE))
# n_all <- length(ci_idx_all)
# 
# # Do NOT wipe previous values — we want to resume
# if (RESET_PREVIOUS) {
#   adaptdata$ClimateImpact_Keywords[ci_idx_all] <- NA_character_
# }
# 
# key_df <- adaptdata |>
#   slice(ci_idx_all) |>
#   transmute(RowID = RowID,
#             Country,
#             Document,
#             ElementText,
#             ElementText_short = make_preview_html(ElementText))
# 
# # ========= Load previous CSV (resume) when requested =========
# if (RESUME_FROM_PREV && file.exists(OUTPUT_CSV)) {
#   prev <- try(read.csv(OUTPUT_CSV, stringsAsFactors = FALSE), silent = TRUE)
#   if (!inherits(prev, "try-error")) {
#     if (!"RowID" %in% names(prev)) {
#       prev <- prev |>
#         rename(ElementText_short = ElementText) |>
#         mutate(RowID = NA_integer_)
#       key_join <- key_df |>
#         left_join(prev[, c("Country","Document","ElementText_short","ClimateImpact_Keywords")],
#                   by = c("Country","Document","ElementText_short"))
#     } else {
#       key_join <- key_df |>
#         left_join(prev[, c("RowID","ClimateImpact_Keywords")], by = "RowID")
#     }
#     filled <- !is.na(key_join$ClimateImpact_Keywords)
#     if (any(filled)) {
#       adaptdata$ClimateImpact_Keywords[match(key_join$RowID[filled], adaptdata$RowID)] <- key_join$ClimateImpact_Keywords[filled]
#       message(sprintf("✔ Loaded %d previously extracted rows from CSV (RESUME_FROM_PREV=TRUE).", sum(filled)))
#     }
#   } else {
#     message("⚠ Could not read existing CSV; continuing without it.")
#   }
# } else if (RESUME_FROM_PREV && !file.exists(OUTPUT_CSV)) {
#   message("⚠ RESUME_FROM_PREV=TRUE but OUTPUT_CSV not found: ", OUTPUT_CSV)
#   # (Optional) auto-pick most recent prior CSV
#   cand <- list.files(OUTPUT_DIR, pattern = "^climate_impact_keywords_.*\\.csv$", full.names = TRUE)
#   if (length(cand)) {
#     OUTPUT_CSV <- cand[which.max(file.info(cand)$mtime)]
#     message("↪ Resuming from: ", basename(OUTPUT_CSV))
#     prev <- try(read.csv(OUTPUT_CSV, stringsAsFactors = FALSE), silent = TRUE)
#     if (!inherits(prev, "try-error")) {
#       if (!"RowID" %in% names(prev)) {
#         prev <- prev |>
#           rename(ElementText_short = ElementText) |>
#           mutate(RowID = NA_integer_)
#         key_join <- key_df |>
#           left_join(prev[, c("Country","Document","ElementText_short","ClimateImpact_Keywords")],
#                     by = c("Country","Document","ElementText_short"))
#       } else {
#         key_join <- key_df |>
#           left_join(prev[, c("RowID","ClimateImpact_Keywords")], by = "RowID")
#       }
#       filled <- !is.na(key_join$ClimateImpact_Keywords)
#       if (any(filled)) {
#         adaptdata$ClimateImpact_Keywords[match(key_join$RowID[filled], adaptdata$RowID)] <- key_join$ClimateImpact_Keywords[filled]
#         message(sprintf("✔ Loaded %d previously extracted rows from auto-selected CSV.", sum(filled)))
#       }
#     }
#   } else {
#     message("⏭ No prior CSVs found; proceeding without resume.")
#   }
# } else {
#   message("⏭ Skipping previous CSV load (RESUME_FROM_PREV=FALSE).")
# }
# 
# # =================== Process remaining ===================
# remaining_rowids <- key_df$RowID[is.na(adaptdata$ClimateImpact_Keywords[match(key_df$RowID, adaptdata$RowID)])]
# n_rem <- length(remaining_rowids)
# message(sprintf("Target rows: %d • Already filled: %d • Remaining: %d",
#                 n_all, n_all - n_rem, n_rem))
# 
# if (n_rem > 0) {
#   use_progress <- requireNamespace("progress", quietly = TRUE)
#   if (use_progress) {
#     pb <- progress::progress_bar$new(
#       format = paste0("Extracting [:bar] :current/:total (:percent) ETA::eta • model=", MODEL, " • last=:last"),
#       total = n_rem, clear = FALSE, width = 80
#     )
#   }
# 
#   for (j in seq_along(remaining_rowids)) {
#     rid <- remaining_rowids[j]
#     i <- match(rid, adaptdata$RowID)
#     txt <- adaptdata$ElementText[i]
# 
#     out <- try(extract_ci_keywords(txt), silent = TRUE)
#     if (inherits(out, "try-error")) {
#       msg <- conditionMessage(attr(out, "condition"))
#       log_error(paste("Unhandled error:", msg))
#       out <- NA_character_
#     }
#     adaptdata$ClimateImpact_Keywords[i] <- out
# 
#     if (use_progress) {
#       pb$tick(tokens = list(last = substr(ifelse(is.na(out), "NA", out), 1, 28)))
#     } else if (j %% 10 == 0) {
#       message(sprintf("... %d/%d done (%.1f%%)", j, n_rem, 100*j/n_rem))
#     }
# 
#     # Autosave checkpoint (atomic) — writes ONLY to this run's CSV
#     if (j %% SAVE_EVERY == 0 || j == n_rem) {
#       ci_preview_all <- adaptdata |>
#         slice(ci_idx_all) |>
#         transmute(RowID,
#                   Country, Document,
#                   ElementText = make_preview_html(ElementText),
#                   ClimateImpact_Keywords)
#       safe_write_csv(ci_preview_all, OUTPUT_CSV)
#     }
#   }
# }
# 
# # =================== Final preview/export ===================
# ci_preview_all <- adaptdata |>
#   slice(ci_idx_all) |>
#   transmute(RowID,
#             Country, Document,
#             ElementText = make_preview_html(ElementText),
#             ClimateImpact_Keywords)
# 
# DT::datatable(
#   ci_preview_all, escape = FALSE,
#   caption = paste0("📄 Climate impact — extracted keywords (", RUN_TAG, ", resume-safe)"),
#   options = list(pageLength = 10, dom = "tip")
# )
# 
# # Final save (to the run-tagged CSV only)
# safe_write_csv(ci_preview_all, OUTPUT_CSV)
# 
# n_na <- sum(is.na(adaptdata$ClimateImpact_Keywords[ci_idx_all]))
# message(sprintf("Done. NA after run: %d of %d target rows.", n_na, n_all))
# if (file.exists(ERROR_LOG)) {
#   message("Some errors were logged to: ", ERROR_LOG)
# }

```


```{r,echo=FALSE}
# ---- Category table for climate impacts (consequences only) ----
suppressPackageStartupMessages({
  library(tibble); library(dplyr); library(stringr); library(purrr); library(readr)
})

impact_categories <- tribble(
  ~Category, ~Keywords, ~Regex,

  "Water availability shortfall",
  "water shortage/shortages/scarcity/stress;water reserves;groundwater levels;fresh water quantities;water levels; reduced availability/flows/groundwater; aquifer depletion; falling levels; increased/rising/higher/growing water demand(s);water management; pressure to meet water demands; supply–demand imbalance; competition for water",
  "(?i)\\b((water (shortage(s)?|scarcity|stress|deficit(s)?))|groundwater supply|water levels|groundwater deficits|(shortage|scarcity) of water|limited (safe )?drinking water (availability|supply)|limited water availability|reduced (availability of )?water|reduced (river|stream|surface water|groundwater|runoff|recharge|discharge)|decreas(?:e|ed|ing) (in )?(river|stream|groundwater|runoff|discharge|flow rates?)|less (river|stream) runoff|\\bwater resource(s)?\\b|\\bwater availability\\b|\\bwater supply\\b|\\bwater resources\\b|lower(?:ing)? (groundwater|water) level(s)?|falling water level(s)?|aquifer (deplet|declin)|water supply (shortage(s)?|deficit(s)?)|(?:increas(?:ed|ing)|rising|higher|growing)\\s+(?:water\\s+)?demand(s)?|pressure to meet water demand(s)?|water (?:demand|consumption) (?:increase|growth|rising)|(?:imbalance|gap) between (?:water )?supply (?:and|&) demand|demand(s)? exceed(?:s|ing) (?:water )?supply|(?:competition|conflict)s? (?:for|over) water)\\b",

  "Water quality deterioration",
  "salinization/saltwater intrusion; contamination; pollutant concentration; microbial/bacterial growth; algal bloom; HABs; hypoxia; nutrient imbalance; turbidity; sedimentation/siltation; reduced/compromised/threatened water quality",
  "(?i)\\b((reduced|diminished) water quality|threaten(?:s|ed|ing)? (drinking )?water quality|compromis(?:e|es|ed|ing) (drinking )?water quality|affect(?:s|ed|ing)? (drinking )?water quality|water pollution|potable water|salini[sz]ation|salt\\s*water intrusion|seawater intrusion|contaminat(?:ed|ion)|pollutant(s)?|higher concentration(s)? of pollutants?|water filtraton|clean water|microbial growth|bacterial growth|algal bloom(s)?|harmful algal bloom(s)?|HABs?|hypox(?:ia|ic)|low oxygen|nutrient imbalance(s)?|eutrophication|turbidity|sediment(?:ation| load|s|ation rates?)|siltation)\\b",

  "Crop production loss",
  "reduced/poor yields; yield decline/variability; crop loss/failure/damage/destroyed;lower yields; reduced agricultural productivity/production;agricultural losses;decreased production;reduce crop yields; postharvest losses;decreased crop yields; affect agricultural production",
  "(?i)\\b(reduced yields?|yield (declin|decreas|variab|loss(es)?)|impacts on crop yields|production loss|production losses|reduce yields|damage to agriculture|impacting agriculture|decrease in food crop production|poor (harvest|yields?)|loss (?:of )?harvest(s)?|harvest (loss|losses)|crop (loss|losses|failure|damage|damages|destroy(?:ed|s)?|destroyed)|damage(?:d)? (?:to )?crops|affect(?:s|ed|ing)? (agricultur\\w* )?production|decreased crop yield(s)|reduce crop yields|reduced (agri\\w* )?(productivity|production)|post[- ]?harvest losses?|impacting productivity|lower agricultural yields?)\\b",

  "Livestock impacts",
  "heat/thermal stress on livestock; livestock disease/mortality; productivity/breeding impacts; feed shortages; cattle loss",
  "(?i)\\b((heat|thermal) stress( (?:in|on))? livestock|livestock (?:disease|loss(?:es)?|livestock health|mortality|deaths?|health issues?|losses in livestock|productivity|livestock diseases|loss of livestock|breeding|breeding (issues|impacts))|feed (shortage|issues)|cattle (killed|loss(?:es)?))\\b",

  "Fisheries & aquaculture decline",
  "fish stock decline; fisheries productivity/catch loss; aquaculture loss; impaired shell formation; coral loss affecting fisheries",
  "(?i)\\b((fisher(?:y|ies)|aquaculture|catch(?:es)?).*(declin|loss|reduct|impact|decreas|fall)|fish stock(s)?\\b|decline in (?:fish )?catch(?: rates)?|catch rates? (declin|drop|fall)|shell (formation|growth) (impair|inhibit))\\b",

  "Food security & nutrition decline",
  "food insecurity/shortage; supply disruptions; hunger; malnutrition/undernutrition; food price spikes/inflation",
  "(?i)\\b(food (insecurity|security (declin|worsen)|shortage|shortages|scarcity|food access|supply (disrupt|disruption))|hunger|grain protein content|malnutrition|under\\s?nutrition|nutrition(?:al)? status (declin|worsen)|food yield instability|food accessibility|food price(s)? (spike|increase|increases|inflation|fluctuation(s)?))\\b",

  "Human health burden",
  "heat-related illness/illnesses/heatstroke;heat-related conditions; respiratory issues; health issues;poor health; allergy; cardiovascular/kidney stress; morbidity;diarrhea; vector-/water-/food-borne disease; diarrhoeal disease;virus; UHI",
  "(?i)\\b(health risks|affect(?:s|ed|ing)? human health|heat[- ]?related illness(?:es)?|heatstroke|heat stress (?:on|for) (?:people|humans|workers)|respiratory (issues|diseases?)|allerg(?:y|ies)|morbidity|epidemic|virus|heat related diseases|respiratory problems|allergenic|poor health|heat stroke|dehydration|mosquito|infectious diseases|health issues|cardiovascular diseases?|kidney (stress|injur(?:y|ies)|failure)|vector[- ]borne diseases?|unhealthy|human lives|water[- ]borne diseases?|threatening the lives|food[- ]borne diseases?|malaria|dengue|Aedes (aegypti|albopictus)|cholera|campylobacter(?:iosis)?|cryptosporid(?:iosis)?|diarr(h)?(ea|heal) (disease|infection|infections)|urban heat island(s)?)\\b",

  "Mental health impacts",
  "anxiety; depression; PTSD; psychological distress; trauma; suicide risk",
  "(?i)\\b(mental (health|well[- ]?being)|psychological (distress|stress)|anxiet(y|ies)|depress(ion|ive)|post[- ]traumatic stress disorder|PTSD|trauma|suicide risk)\\b",

  "Healthcare system strain",
  "overburdened health services;health infrastructures;healthcare system; hospital admissions rising; overwhelmed/closed clinics/hospitals; service backlogs",
  "(?i)\\b((health\\s*care|healthcare) (system|services?).*(overburdened|strained|overwhelmed|capacity (exceed|shortage)|backlog(s)?)|increase(d)? hospital (admission|admissions)|overburdened health services|(?:clinic|hospital)(s)? (overwhelmed|nonfunctional|closed)|shortages of medical supplies|healthcare facilities|healthcare system|public health)\\b",

  "Mortality & injury",
  "deaths/fatalities/mortality;died; loss of life; people killed; drowning; injuries; casualties; excess deaths",
  "(?i)\\b(deaths?|fatalit(?:y|ies)|mortality|loss(?:es)? of life|people (?:killed|dead)|\\d+ (?:people|persons) (?:killed|dead)|drown(?:ing|ed)|injur(?:y|ies|ed)|died|loss of human life|casualt(?:y|ies)|excess deaths?)\\b",

  "Displacement & migration",
  "displacement; evacuation; relocation; forced migration",
  "(?i)\\b(displac(?:e|ed|ement)|evacuat(?:e|ed|ion)|relocat(?:e|ed|ion)|forced migration|migrants?)\\b",

 "Damage to buildings & infrastructure",
"houses/buildings damaged or destroyed; property/infrastructure damage; shortened asset lifespan; reservoir damage",
"(?i)\\b(
   (house(s)?|home(s)?|building(s)?|property|infrastructure|reservoir(s)?).*(damage(?:d)?|destroy(?:ed|s)?|deteriorat|loss(?:es)?) |aging infrastructure|damage to infrastructure|damage to buildings|buildings|
   (damage(?:d)?|destroy(?:ed|s)?|deteriorat|loss(?:es)?).*(house(s)?|home(s)?|building(s)?|property|infrastructure|reservoir(s)?) |damaging infrastructure|inadequate infrastructure|lack proper infrastructure|endanger infrastructure|
   (lifespan|service life) (reduc|shorten|decreas)(ed)?
)\\b
|\\b(houses|buildings)\\s+damaged\\s+or\\s+destroyed\\b
|\\b(property|infrastructure)\\s+damage\\b
|\\bshortened\\s+asset\\s+lifespan\\b
|\\breservoir\\s+damage\\b",


  "Energy system impacts",
  "reduced hydropower; outages/power cuts; grid/transmission issues; reduced plant efficiency; increased cooling demand; reduced heating demand; coal quality reduced",
  "(?i)\\b(hydropower|power cut(s)?|outage(s)?|electricit(?:y|y grid)|substation(s)?|energy|cooling|transmission (line|lines|efficiency)|grid|plant efficiency|coal quality|energy demand (for )?cooling|cooling demand|reduced heating demand|energy (supply|system))\\b",

  "Transport disruption",
  "roads/bridges/rail/ports/airports damaged; access routes blocked/cut/closed; services disrupted; road damage/closures",
  "(?i)\\b((road(s)?|bridge(s)?|rail(way)?|port(s)?|airport(s)?).*(damaged|destroyed|washed (?:away|out)|closed|closure|affected|disrupted)|road (damage|closures?)|rail(way)? (damage|closures?)|affect accessibility|transportroads|access route(s)? (blocked|cut|closed)|road|traffic|railway|transport infrastructure|transport (services?|systems?) (disrupt|hinder|affect|suspended))|ports|highways\\b",

  "Supply chain & logistics disruptions",
  "supply chains disrupted; logistics delays; port operations disrupted/suspended; access routes cut",
  "(?i)\\b(supply chain(s)? (disrupt(?:ed|ion)|delay(s)?|interrupt(?:ed|ion))|logistics (disrupt(?:ed|ion)|delay(s)?)|port (operation|operations) (disrupt(?:ed|ion)|suspended)|access route(s)? (blocked|cut|closed)|supply chain)\\b",

  "Economic & livelihood losses",
  "GDP loss; economic losses/costs; productivity loss; labor productivity decline; livelihood/income/job loss; losses & damages; operating/compliance costs; price spikes",
  "(?i)\\b(GDP\\b|economic (loss|losses|damage|damages|costs)|destroyed several farm|socio-economic impact|rural livelihoods|property losses|productivity (loss|declin|reduct)|decreas(?:e|ed|ing) (labor|worker) productivity|livelihood(s)? (loss|impact)|homes damaged|income (loss|declin)|job(s)? (loss|cuts)|financial losses|losses? and damages?|operational cost(s)? (increase|increases)|property damage|employment|job|jobs|livelihoods|poor housing conditions|food prices|damage to property|precarious location|local economies|compliance cost(s)? (increase|increases)|price (spike|spikes|increase|increases|fluctuation(s)?))\\b",

  "Tourism & recreation impacts",
  "reduced tourist numbers; shorter/shortened snow season; coastal tourism loss; tourism revenue decline; visitors discouraged;outdoor activities; shorter stays; outdoor activities hindered;reduce tourist numbers;shorten average stays",
  "(?i)\\b((?:touris(m|t)s?|visitor(s)?|recreation|ski|tourism|winter tourism|reduced tourist arrivals|negatively impact tourism|affect tourism|tourist|snow cover|beach(es)?|winter (season|touris(m|t))).*(reduced|declin|shorter|shorten(?:ed|ing)?|loss|damage|discourage|deter|unviable|impact)|tourism revenue|tourism demand|average (stay|stays) (shorten|shorter)|outdoor activities? (hinder(?:ed|ing)?|limited|snow cover|tourist))\\b",

  "Ecosystem degradation & biodiversity loss",
  "habitat degradation/loss; ecosystem integrity/functions reduced; biodiversity loss; loss of forest functions; species decline/shift/extinction;coral bleaching;loss of habitats; invasive species;forest degradation; disruption of wildlife;habitat degradation;loss of species; habitat deterioration;ecosystem service loss; reduce ecosystem service; forest health/growth decline; wildfire severity increase; coral loss/bleaching; wildlife disruption",
  "(?i)\\b(habitat (degrad|deteriorat|loss)|biodiversit(?:y|y loss)|reef loss|coral cover|tree diversity|loss of species|key species|ecosystem (integrity|function(s)?|services?) (declin|loss|disrupt|reduct)|disrupt(?:ion of )?ecosystem(s)?|wildlife (disrupt|disturb|declin)|disruption of wildlife|genetic diversity|habitat deterioration|habitat degradation|habitat|habitats|deforestation|distribution ranges|extinction|extinctions|species (declin|loss|extinct|range (shift|reduc))|invasive species|ecosystem (service(s)? (loss|reduct)|degrad|function(s)? (disrupt|loss))|forest (health|degrad|dieback|mortality)|wildfire (severity|risk) (increase|increased)|coral (reef )?(loss|degrad|bleach)|tree (growth|health) (declin|reduc))\\b",

  "Soil erosion & land degradation",
  "soil erosion; land degradation; desertification; loss of soil fertility/quality; gully erosion; topsoil loss",
  "(?i)\\b(soil erosion|land degradation|runoff|runoffs|desertification|gully erosion|loss of soil (fertility|quality)|topsoil (loss|deplet(?:ion)?))\\b",

  "Marine ecosystem stressors",
  "marine heatwaves; ocean acidification; hypoxia/low oxygen; red/green tides; jellyfish blooms; nutrient imbalance",
  "(?i)\\b(marine heatwave(s)?|ocean acidification|hypox(?:ia|ic)|low oxygen|red tide(s)?|green tide(s)?|jellyfish bloom(s)?|nutrient imbalance(s)?)\\b",

  "Cultural heritage impacts",
  "cultural heritage/values damaged/destroyed",
  "(?i)\\b(cultural[- ]?heritage|cultural[- ]?historical values|heritage site(s)?|monument(s)?).*(damage|destroy|degrad|loss)\\b",
  
"Service delivery disruptions",
"water/electricity/sanitation/waste/telecom services disrupted/failing; cold chain affected",
"(?i)\\b(
    service(s)? (disrupt|interruption) |
    water service(s)? (disrupt|interruption) |
    electricit(?:y|y services?) (disrupt|interruption) |
    sanitation (service|system)s? (disrupt|failure) |
    cold chain (storage|facilit(?:y|ies)) (impacted|disrupted) |telecommunications|
    waste management (system|services?) (impacted|disrupted) |energy infrastructure|
    (telecommunication(s)?|broadband|internet|ICT|communication network(s)?) (disrupt|failure)
)\\b",


  "Carbon storage & productivity decline",
  "NPP decline; reduced net primary productivity; soil carbon stocks decline; lower sequestration",
  "(?i)\\b((net primary productivity|NPP) (declin|decreas|reduc)|soil carbon stocks? (declin|decreas|reduc)|carbon sequestration (declin|decreas|reduc))\\b",

  "Coastal inundation & shoreline erosion",
  "coastal erosion/retreat; shoreline retreat; permanent inundation; coastal/storm-surge flooding; land loss to sea; tidal/nuisance flooding; below the high tide line",
  "(?i)\\b(coastal (erosion|retreat|inundation|flood(ing)?)|shoreline (retreat|erosion)|beach (erosion|loss)|land loss (to|from) the sea|sea level|level of the sea|sea levels|coastal|sea level rise|permanent inundation|storm surge (inundation|flood(ing)?)|tidal flooding|nuisance flooding|below the high tide line)\\b",

  "Cryosphere degradation & ground instability",
  "permafrost thaw/degradation; active layer thickening; thermokarst; ground subsidence/instability; frost heave; glacier melt/retreat; reduced snow cover",
  "(?i)\\b(permafrost (thaw|degrad|melt)|active layer (thicken|deepening)|glacier|ice|thermokarst|permafrost|ground (subsidence|instability)|frost heave|glacier (melt|retreat|shrinkage)|snow cover (declin|reduc|loss))\\b",


  "Air quality deterioration",
  "air pollution; poor air quality; PM2.5/PM10; particulate matter; ozone/smog; smoke; haze; dust storms",
  "(?i)\\b(air (pollution|quality (declin|deteriorat|worsen))|PM\\s?(2\\.5|10)\\b|particulate matter|ground[- ]?level ozone|ozone\\b|smog|greenhouse gas emissions|pollution|smoke|haze|dust storm(s)?)\\b",

  "Agricultural pests & diseases",
  "crop pest/disease outbreaks;pest infestations;pests;alien species pathogen pressure; spread of pests; pest incidence; incidence/severity/spread; overwintering pests; higher pesticide use",
  "(?i)\\b((crop|agricultur\\w*) (pest(s)?|disease(s)?|pathogen(s)?)( outbreak(s)?| incidence| incidences| occurrence(s)?| pressure| severity| spread)|overwintering (pests|insects)|alien species|pest|pests|higher pesticide use|pest|pests|pesticide (use|usage) (increase|rising|higher))\\b",

  "Agricultural system shifts",
  "growing season lengthened/shortened; earlier green-up; delayed senescence; planting boundary shifting northward/poleward/higher elevations;phenology; multi-cropping index rising;shift to higher elevations;shifting in agro-ecological zones;shifting in production zones;frost damage; cropping systems shifting",
  "(?i)\\b(growing season (lengthen(?:ed|ing)?|shorten(?:ed|ing)?)|geographic distribution|shift to higher elevations|shifting in agro-ecological zones|shift|shifting|shift in production zones|phenological|earlier green[- ]?up|delayed senescence|planting boundary (shift(?:ing)?|moving) (northward|poleward|to higher (latitudes|altitudes|elevations))|multi-?cropping (index|area(s)?) (increase|rising|expanded)|cropping system(s)? (shift|change))\\b",

"Potential gains",
"longer/extended growing season; increased crop yields/productivity/production; increased timber; increased NPP/biomass;attractive; multi-cropping index rising;increase crop yields;positive effect;beneficial;reduce heating demand; new species compositions improving services; reduced heating demand",
"(?i)\\b((longer|extended|lengthen(?:ed|ing))\\s+growing\\s+season(s)?|increase(?:d|s|ing)\\s+(?:crop\\s+)?(?:yield(s)?|productivity|production)\\b(?![^.]{0,20}\\b(decreas|declin|loss|reduct))|increase(?:d|s|ing)\\s+(?:NPP|biomass|forest\\s+productivity|ecosystem\\s+productivity)\\b(?![^.]{0,20}\\b(decreas|declin|loss|reduct))|multi-?cropping\\s+index(es)?\\s+(?:increase|rising|higher)|new\\s+species\\s+composition(s)?\\s+(?:improv(?:e|ing)|enhanc(?:e|ing))\\s+ecosystem\\s+services|reduced\\s+heating\\s+demand|increase\\s+crop\\s+yields?|increase\\s+in\\s+the\\s+performance|increased\\s+availability\\s+of\\s+feed|expand\\s+the\\s+areas\\s+suitable\\s+for\\s+agriculture|increasing\\s+productivity|expansion\\s+of\\s+the\\s+crop\\s+production|increasing\\s+efficiency|facilitating\\s+access|reduction\\s+of\\s+energy\\s+consumption)\\b"
)

# --- Ensure every Keywords phrase is covered by the Regex for its category ---

escape_regex <- function(x) str_replace_all(x, "([\\.^$|?*+()\\[\\]{}])", "\\\\\\1")
hyphen_flexible <- function(x) str_replace_all(x, "\\s*[–-]\\s*", "\\\\s*[–-]\\\\s*")

token_to_pattern <- function(token) {
  token <- str_squish(token)
  if (token == "") return("")
  alts <- str_split(token, "/", simplify = TRUE) %>% as.character()
  alts <- escape_regex(alts) %>% hyphen_flexible()
  if (length(alts) == 1) alts else paste0("(?:", paste(alts, collapse="|"), ")")
}

phrase_to_regex <- function(phrase) {
  tokens <- str_split(str_squish(phrase), "\\s+", simplify = TRUE) %>% as.character()
  if (!length(tokens)) return("")
  pats <- map_chr(tokens, token_to_pattern)
  paste0("\\\\b", paste(pats, collapse="\\\\s+"), "\\\\b")
}

keywords_to_regex <- function(keywords) {
  phrases <- str_split(keywords, ";", simplify = TRUE) %>% as.character() %>% str_trim()
  phrases <- phrases[nzchar(phrases)]
  if (!length(phrases)) return("")
  paste(map_chr(phrases, phrase_to_regex), collapse="|")
}

strip_leading_caseflag <- function(rx) rx %>% str_replace("^\\(\\?i\\)", "") %>% str_squish()

ensure_keywords_covered <- function(tbl) {
  tbl %>%
    mutate(
      .kw_regex = map_chr(Keywords, keywords_to_regex),
      .existing = map_chr(Regex, strip_leading_caseflag),
      Regex = pmap_chr(list(.existing, .kw_regex), function(exist, kw) {
        if (!nzchar(exist) && !nzchar(kw)) return("(?i)")
        if (!nzchar(exist)) return(paste0("(?i)(?:", kw, ")"))
        if (!nzchar(kw))   return(paste0("(?i)", exist))
        paste0("(?i)(?:", exist, "|", kw, ")")
      })
    ) %>%
    select(-.kw_regex, -.existing)
}

impact_categories <- ensure_keywords_covered(impact_categories)

# ---- Matching helpers & pipeline ----

# Read your extracted keywords file
adaptdata <- read_csv("adaptdata_climateimpact5nano.csv")

# Sanity check
stopifnot(exists("impact_categories"),
          all(c("Category","Regex","Keywords") %in% names(impact_categories)))

clean_ci_text <- function(x) {
  x %>% str_replace_all("<[^>]+>", " ") %>% str_squish()
}

# Optionally drop rows that only list hazards (no consequences)
hazard_only_regex <- "(?i)\\b(drought(s)?|flood(s|ing)?|storm(s)?|storm surge(s)?|cyclone(s)?|typhoon(s)?|hurricane(s)?|heatwave(s)?|heat wave(s)?|heavy rainfall|rainstorm(s)?|sea[ -]?level rise|landslide(s)?|wildfire(s)?|windstorm(s)?)\\b"

match_categories <- function(txt, map = impact_categories, drop_hazard_only = TRUE) {
  if (is.na(txt) || !nzchar(txt)) return(character(0))
  s <- clean_ci_text(txt)

  if (drop_hazard_only) {
    any_consequence <- any(str_detect(s, map$Regex))
    only_hazard <- (!any_consequence) && str_detect(s, hazard_only_regex)
    if (only_hazard) return(character(0))
  }

  hits <- map$Category[ map_lgl(map$Regex, ~ str_detect(s, .x)) ]
  unique(hits)
}

categorize_ci_vector <- function(v, map = impact_categories, drop_hazard_only = TRUE) {
  map_chr(v, ~{
    cats <- match_categories(.x, map = map, drop_hazard_only = drop_hazard_only)
    if (length(cats) == 0) NA_character_ else paste(cats, collapse = "; ")
  })
}

# ---- Create ONE column with semicolon-separated categories ----
adaptdata <- adaptdata %>%
  mutate(
    ImpactCategories = categorize_ci_vector(ClimateImpact_Keywords,
                                            map = impact_categories,
                                            drop_hazard_only = TRUE)
  )

# Coverage metric (fixed object name)
coverage <- mean(!is.na(adaptdata$ImpactCategories))
message(sprintf("Impact categorization coverage: %.1f %%", 100 * coverage))

# Optional: write out the categorized file
# write_csv(adaptdata, "adaptdata_climate_impact_categorized.csv")

# Quick export for spot-checking
write.csv(adaptdata, "adaptdata_climateimpact5nano.csv", row.names = FALSE)


```

## 4.3 Climate impact summary data 

```{r,echo=FALSE}

# 1. Filter and separate multiple SystemType values
impact_counts <- adaptdata %>%
  filter(Element == "Climate impact",
         !is.na(ImpactCategories),
         str_detect(ImpactCategories, "\\S")) %>%            # has any non-space
  separate_rows(ImpactCategories, sep = ";") %>%             # split on ;
  mutate(
    ImpactCategories = str_replace_all(ImpactCategories, "\\p{Zs}+", " "), # normalize unicode spaces
    ImpactCategories = str_squish(ImpactCategories)                         # trim ends + collapse internal
  ) %>%
  filter(ImpactCategories != "") %>%                          # drop empty tokens
  count(ImpactCategories, sort = TRUE) %>%
  rename(`Impact Category` = ImpactCategories,
         Occurrences = n)

# 2. Display the interactive table
DT::datatable(
  impact_counts,
  rownames = FALSE,
  options = list(
    pageLength = 10,
    autoWidth = TRUE,
    dom = 'tip'
  )
)

write.csv(impact_counts,   "impact_counts.csv", row.names = FALSE)

# 3. Visualize systems at risk
system_plot <- ggplot(impact_counts,
       aes(x = reorder(`Impact Category`, Occurrences),
           y = Occurrences,
           fill = `Impact Category`)) +
  geom_col(show.legend = FALSE) +
  coord_flip() +
  scale_fill_viridis_d(option = "mako") +
  labs(title = "Climate impacts by count", x = NULL, y = "Occurrences") +
  theme_minimal(base_size = 12) +
  theme(
    axis.text.y = element_text(size = 14),
    axis.text.x = element_text(size = 14)
  )

system_plot


```

# Step 5 : Action level and type
## 5.1 Action level

## Step X: Action Level classification

**Goal.** Distinguish whether each *Action* reported is **preparatory groundwork** or a **substantive adaptation measure**.

**Definitions**

- **Groundwork** — upstream or enabling work such as:
  - impact/vulnerability assessments, scenarios, conceptual tools  
  - policy recommendations, planning frameworks  
- **Substantive action** — concrete measures such as:
  - organizational development, regulations, awareness/outreach, education/training  
  - monitoring/MEL, infrastructure, technology/innovation  
  - financial mechanisms, resource transfer, funding  

**Method**

- GPT reads each *Action* text and must return **exactly one label**: *Groundwork* or *Substantive action*.  
- If no clear match, the model returns `NA`.  
- The result is stored in `ActionLevel_GPT` and previewed in a datatable.  



```{r,echo=FALSE}
# # ====================== Step X: Action Level (GPT) ======================
# # Classify Element == "Action" into ONE of:
# #   Groundwork | Substantive action
# # Uses ellmer::chat_openai ONLY and BTR_KEY for auth.
# 
# suppressPackageStartupMessages({
#   library(dplyr)
#   library(stringr)
#   library(glue)
#   library(DT)
#   library(ellmer)
# })
# 
# # ---- API key: force BTR_KEY for this session ----
# btr_key <- Sys.getenv("BTR_KEY")
# stopifnot("BTR_KEY is not set in your environment." = nzchar(btr_key))
# Sys.setenv(OPENAI_API_KEY = btr_key)  # ellmer reads this
# 
# # ---- Model ----
# ACT_LEVEL_MODEL <- "gpt-5-nano"
# 
# # ---- Allowed labels ----
# ACTION_LEVEL_LABELS <- c("Groundwork", "Substantive action")
# 
# # ---- Prompt (definitions + strict output) ----
# make_action_level_prompt <- function(txt) {
#   glue::glue(
# "Decide the *Action level* for the passage. Choose ONE label only.
# 
# Allowed labels:
# - Groundwork — preparatory work (e.g., scenarios, impact/vulnerability assessment, conceptual tools, policy recommendations).
# - Substantive action — concrete adaptation actions (e.g., organizational development, regulations, public and raising awareness/outreach/capacity building/education, surveillance/monitoring/MEL, infrastructure/technology/innovation, resource transfer/financial mechanisms,funding, financing or co financing).
# 
# Rules:
# - Think silently if the action is concrete or abstract; do NOT show your reasoning.
# - Output exactly ONE label from the list above.
# - If none fit, return 'NA'.
# 
# TEXT:
# \"\"\"{txt}\"\"\""
#   )
# }
# 
# # ---- Single-call classifier (ONE label) ----
# classify_action_level <- function(txt, model = ACT_LEVEL_MODEL) {
#   if (is.na(txt) || !nzchar(txt)) return(NA_character_)
#   ans <- tryCatch(
#     ellmer::chat_openai(model = model)$chat(make_action_level_prompt(txt)),
#     error = function(e) e
#   )
#   if (inherits(ans, "error")) {
#     warning("GPT call failed: ", conditionMessage(ans))
#     return(NA_character_)
#   }
#   raw <- trimws(ans)
#   if (!nzchar(raw) || toupper(raw) == "NA") return(NA_character_)
#   # Canonicalize to allowed label (case-insensitive)
#   hit <- ACTION_LEVEL_LABELS[tolower(ACTION_LEVEL_LABELS) == tolower(raw)]
#   if (!length(hit)) return(NA_character_)
#   hit
# }
# 
# # ---- Run on Action rows ----
# if (!"ActionLevel_GPT" %in% names(adaptdata)) adaptdata$ActionLevel_GPT <- NA_character_
# 
# act_idx <- which(grepl("^\\s*action\\s*$", adaptdata$Element, ignore.case = TRUE))
# if (length(act_idx)) {
#   cat("Classifying Action Level for", length(act_idx), "rows...\n")
#   pb <- txtProgressBar(min = 0, max = length(act_idx), style = 3)
#   for (i in seq_along(act_idx)) {
#     r <- act_idx[i]
#     adaptdata$ActionLevel_GPT[r] <- classify_action_level(adaptdata$ElementText[r])
#     setTxtProgressBar(pb, i)
#   }
#   close(pb)
# } else {
#   cat("No rows with Element == 'Action' found.\n")
# }
# 
# # ---- Preview ----
# preview_level <- adaptdata %>%
#   filter(grepl("^\\s*action\\s*$", Element, ignore.case = TRUE)) %>%
#   transmute(
#     Country, Document,
#     ElementText = ifelse(is.na(ElementText), "",
#                          ifelse(nchar(ElementText) > 120,
#                                 paste0("<span title='", ElementText, "'>", substr(ElementText, 1, 120), "...</span>"),
#                                 ElementText)),
#     ActionLevel_GPT
#   )
# 
# DT::datatable(
#   preview_level, escape = FALSE,
#   caption = "🧭 Action level (Groundwork vs Substantive) — GPT",
#   options = list(pageLength = 10, autoWidth = TRUE, dom = "tip")
# )
# 
# # Optional export
# write.csv(preview_level, "action_level_gpt_preview.csv", row.names = FALSE)
# write.csv(adaptdata,"adaptdata_action_level.csv")
```

## 5.2 Action type

**Goal.** Tag each *Action* with a single **type** to describe the primary adaptation approach.

**Types (one label only)**

- **Infrastructural/ technological** — building or upgrading assets and systems; deploying tech or tools.  
  *E.g., flood defenses, retrofits, sensors, drought-resistant irrigation.*
- **Behavioral/ cultural** — influencing practices and norms.  
  *E.g., public awareness, training, behavior change campaigns.*
- **Institutional** — strengthening rules, mandates, coordination, and org capacity.  
  *E.g., policies, plans, regulatory updates, new units/committees.*
- **Ecosystem-based** — conserving, restoring, or using ecosystems for risk reduction.  
  *E.g., mangrove restoration, watershed rehab, green infrastructure.*
- **Economic/financial** — funding flows and incentives.  
  *E.g., subsidies, insurance, green bonds, grants, resource transfers.*

**Method**

- GPT reads the *Action* text and must return **exactly one** of the five labels (or `NA` if unclear).
- The output is saved in `ActionType_GPT` and previewed in a datatable.


```{r,echo=FALSE}
# # ====================== Step Y: Action Type (GPT) ======================
# # Classify Element == "Action" into ONE of:
# #   Infrastructural/ technological | Behavioral/ cultural | Institutional | Ecosystem-based | Economic/financial
adaptdata<- read_csv("adaptdata_action_level.csv")
# 
# suppressPackageStartupMessages({
#   library(dplyr)
#   library(stringr)
#   library(glue)
#   library(DT)
#   library(ellmer)
# })
# 
# # ---- API key: force BTR_KEY again (safe if repeated) ----
# btr_key <- Sys.getenv("BTR_KEY")
# stopifnot("BTR_KEY is not set in your environment." = nzchar(btr_key))
# Sys.setenv(OPENAI_API_KEY = btr_key)
# 
# # ---- Model ----
# ACT_TYPE_MODEL <- "gpt-5-nano"
# 
# # ---- Allowed labels ----
# ACTION_TYPE_LABELS <- c(
#   "Infrastructural/ technological",
#   "Behavioral/ cultural",
#   "Institutional",
#   "Ecosystem-based",
#   "Economic/financial"
# )
# 
# # ---- Prompt (definitions + strict output) ----
# make_action_type_prompt <- function(txt) {
#   glue::glue(
# "Decide the *Action type* for the passage. Choose ONE label only.
# 
# Allowed labels:
# - Infrastructural/ technological — enabling or implementing technology or infrastructure development.
# - Behavioral/ cultural — enabling or implementing lifestyle/behavioural change.
# - Institutional — improving the institutional/organizational context (policies, plans, org structures).
# - Ecosystem-based — enhancing/protecting/promoting ecosystem services and products.
# - Economic/financial — finance instruments, incentives, resource transfers, funding mechanisms.
# 
# Rules:
# - Think silently; do NOT show your reasoning.
# - Output exactly ONE label from the list above.
# - If none fit, return 'NA'.
# 
# TEXT:
# \"\"\"{txt}\"\"\""
#   )
# }
# 
# # ---- Single-call classifier (ONE label) ----
# classify_action_type <- function(txt, model = ACT_TYPE_MODEL) {
#   if (is.na(txt) || !nzchar(txt)) return(NA_character_)
#   ans <- tryCatch(
#     ellmer::chat_openai(model = model)$chat(make_action_type_prompt(txt)),
#     error = function(e) e
#   )
#   if (inherits(ans, "error")) {
#     warning("GPT call failed: ", conditionMessage(ans))
#     return(NA_character_)
#   }
#   raw <- trimws(ans)
#   if (!nzchar(raw) || toupper(raw) == "NA") return(NA_character_)
#   # Canonicalize to allowed label (case-insensitive)
#   hit <- ACTION_TYPE_LABELS[tolower(ACTION_TYPE_LABELS) == tolower(raw)]
#   if (!length(hit)) return(NA_character_)
#   hit
# }
# 
# # ---- Run on Action rows ----
# if (!"ActionType_GPT" %in% names(adaptdata)) adaptdata$ActionType_GPT <- NA_character_
# 
# act_idx <- which(grepl("^\\s*action\\s*$", adaptdata$Element, ignore.case = TRUE))
# if (length(act_idx)) {
#   cat("Classifying Action Type for", length(act_idx), "rows...\n")
#   pb <- txtProgressBar(min = 0, max = length(act_idx), style = 3)
#   for (i in seq_along(act_idx)) {
#     r <- act_idx[i]
#     adaptdata$ActionType_GPT[r] <- classify_action_type(adaptdata$ElementText[r])
#     setTxtProgressBar(pb, i)
#   }
#   close(pb)
# } else {
#   cat("No rows with Element == 'Action' found.\n")
# }
# 
# # ---- Preview ----
# preview_type <- adaptdata %>%
#   filter(grepl("^\\s*action\\s*$", Element, ignore.case = TRUE)) %>%
#   transmute(
#     Country, Document,
#     ElementText = ifelse(is.na(ElementText), "",
#                          ifelse(nchar(ElementText) > 120,
#                                 paste0("<span title='", ElementText, "'>", substr(ElementText, 1, 120), "...</span>"),
#                                 ElementText)),
#     ActionType_GPT
#   )
# 
# DT::datatable(
#   preview_type, escape = FALSE,
#   caption = "🏷️ Action type — GPT",
#   options = list(pageLength = 10, autoWidth = TRUE, dom = "tip")
# )
# 
# # Optional export
# write.csv(preview_type, "action_type_gpt_preview.csv", row.names = FALSE)
# write.csv(adaptdata,"adaptdata_action_type.csv")

```

```{r,echo=FALSE}

# --- Action Level & Type summaries (counts + DT + plots) --------------------
suppressPackageStartupMessages({
  library(dplyr); library(stringr); library(tidyr)
  library(DT); library(ggplot2); library(readr)
})

# If the current environment doesn't have both columns, try reloading the latest CSV
if (!exists("adaptdata") || !all(c("ActionLevel_GPT","ActionType_GPT") %in% names(adaptdata))) {
  if (file.exists("adaptdata_action_type.csv")) {
    adaptdata <- readr::read_csv("adaptdata_action_type.csv", show_col_types = FALSE)
  } else if (file.exists("adaptdata_action_level.csv")) {
    adaptdata <- readr::read_csv("adaptdata_action_level.csv", show_col_types = FALSE)
  } else {
    stop("adaptdata with ActionLevel_GPT / ActionType_GPT not found. Run steps 5.1–5.2 first.")
  }
}

# Helper to normalize unicode spaces and trim tokens
.norm_tok <- function(x) {
  x %>% str_replace_all("\\p{Zs}+", " ") %>% str_squish()
}

# -----------------------
# A) ACTION LEVEL SUMMARY
# -----------------------
level_counts <- adaptdata %>%
  filter(grepl("^\\s*action\\s*$", Element, ignore.case = TRUE),
         !is.na(ActionLevel_GPT),
         str_detect(ActionLevel_GPT, "\\S")) %>%
  separate_rows(ActionLevel_GPT, sep = ";") %>%
  mutate(ActionLevel_GPT = .norm_tok(ActionLevel_GPT)) %>%
  filter(ActionLevel_GPT != "") %>%
  count(ActionLevel_GPT, sort = TRUE) %>%
  rename(`Action Level` = ActionLevel_GPT,
         Occurrences = n)

DT::datatable(
  level_counts,
  rownames = FALSE,
  options = list(pageLength = 10, autoWidth = TRUE, dom = 'tip'),
  caption = "🧭 Action Level (Groundwork vs Substantive) — counts"
)

level_props <- level_counts %>%
  mutate(Percent = Occurrences / sum(Occurrences),
         midlab  = cumsum(Percent) - Percent/2,
         label   = paste0(round(Percent*100), "%"))

donut <- ggplot(level_props, aes(x = 2, y = Occurrences, fill = `Action Level`)) +
  geom_col(width = 1, color = NA) +
  # inner white circle to make the "hole"
  geom_col(data = data.frame(Occurrences = sum(level_props$Occurrences)),
           aes(x = 1, y = Occurrences),
           inherit.aes = FALSE, fill = "white", width = 1) +
  coord_polar(theta = "y") +
  xlim(0.5, 2.5) +
  scale_fill_viridis_d(option = "viridis") +
  labs(title = "Action level — composition") +
  theme_void(base_size = 12) +
  theme(legend.title = element_blank())

donut



# ----------------------
# B) ACTION TYPE SUMMARY
# ----------------------
type_counts <- adaptdata %>%
  filter(grepl("^\\s*action\\s*$", Element, ignore.case = TRUE),
         !is.na(ActionType_GPT),
         str_detect(ActionType_GPT, "\\S")) %>%
  separate_rows(ActionType_GPT, sep = ";") %>%        # robust if multiple
  mutate(ActionType_GPT = .norm_tok(ActionType_GPT)) %>%
  filter(ActionType_GPT != "") %>%
  count(ActionType_GPT, sort = TRUE) %>%
  rename(`Action Type` = ActionType_GPT,
         Occurrences = n)

DT::datatable(
  type_counts,
  rownames = FALSE,
  options = list(pageLength = 10, autoWidth = TRUE, dom = 'tip'),
  caption = "🏷️ Action Type — counts"
)

write.csv(type_counts, "action_type_counts.csv", row.names = FALSE)

type_plot <- ggplot(type_counts,
  aes(x = reorder(`Action Type`, Occurrences),
      y = Occurrences,
      fill = `Action Type`)) +
  geom_col(show.legend = FALSE) +
  coord_flip() +
  scale_fill_viridis_d(option = "mako") +
  labs(title = "Action Type by count", x = NULL, y = "Occurrences") +
  theme_minimal(base_size = 12) +
  theme(axis.text.y = element_text(size = 14),
        axis.text.x = element_text(size = 14))

type_plot

```

# Step 6 : Results

**Goal.** Label each *Result* as one of four types so we can compare what’s being reported across countries and sectors.

**Types (choose one)**  
- **Input** — resources to implement actions (funding, staff, tools, data, analyses, maps).  
- **Process** — planning/governance steps (plans, strategies, frameworks, integration into policy).  
- **Output** — immediate deliverables (things built/distributed, trainings, coverage/area).  
- **Outcome** — condition/performance changes (yields, resilience, wellbeing, recovery).

**Method**  
- GPT reads the *Result* text and returns **exactly one label** from the list above (or `NA` if unclear).  
- We **store the model’s reply verbatim** in `ResultType_GPT` (no normalization here); any cleanup or remapping happens later.


```{r,echo=FALSE}

adaptdata <- read_csv("adaptdata_action_type.csv")
# ========================= Step X: Result Type (GPT) =========================
# Goal now: store EXACT GPT output in ResultType_GPT (no normalization/cleanup)
# 
# suppressPackageStartupMessages({
#   library(dplyr)
#   library(stringr)
#   library(glue)
#   library(DT)
#   library(ellmer)
# })
# 
# # ---- API key: FORCE using BTR_KEY this session ----
# btr_key <- Sys.getenv("BTR_KEY")
# stopifnot("BTR_KEY is not set in your environment." = nzchar(btr_key))
# Sys.setenv(OPENAI_API_KEY = btr_key)  # ellmer reads OPENAI_API_KEY
# 
# # ---- Model ----
# RESULT_MODEL <- "gpt-5-nano"
# 
# # ---- Prompt builder (unchanged, but we will store raw model output) ----
# make_result_prompt <- function(txt) {
#   glue::glue(
# "Classify the passage by the *type of results* it reports. You may choose one or more.
# 
# Allowed labels (choose from these EXACT words only):
# - Input — resources needed to implement an action (money,funding, staff, tools, data, analyses, maps).
# - Process — policies/steps/governance to plan or implement (plans, strategies, frameworks, integration into policies).
# - Output — immediate products/services delivered (things built, distributed, trained people, area covered).
# - Outcome — changes in condition or performance (improved yields, efficiency, resilience, recovery, wellbeing).
# 
# Rules:
# - Think silently; DO NOT write your reasoning.
# - if more than one category could fit choose the most relevant one for the description.
# - Return ONLY the label you choose.
# - No extra words or punctuation. If none fit, return 'NA'.
# 
# TEXT:
# \"\"\"{txt}\"\"\""
#   )
# }
# 
# # ---- Raw classifier: returns EXACT text from GPT (no trimming/canonization) ----
# classify_result_types_raw <- function(txt, model = RESULT_MODEL) {
#   if (is.na(txt) || !nzchar(txt)) return(NA_character_)
#   prompt <- make_result_prompt(txt)
# 
#   ans <- tryCatch(
#     ellmer::chat_openai(model = model)$chat(prompt),
#     error = function(e) e
#   )
#   if (inherits(ans, "error")) {
#     warning("GPT call failed: ", conditionMessage(ans))
#     return(NA_character_)
#   }
# 
#   # Store *exactly* what the model returned
#   as.character(ans)
# }
# 
# # ---- Run on Element == "Result" (store raw output) ----
# if (!"ResultType_GPT" %in% names(adaptdata)) adaptdata$ResultType_GPT <- NA_character_
# 
# res_idx <- which(grepl("^\\s*result\\s*$", adaptdata$Element, ignore.case = TRUE))
# 
# if (length(res_idx)) {
#   cat("Classifying", length(res_idx), "rows where Element == 'Result'...\n")
#   pb <- txtProgressBar(min = 0, max = length(res_idx), style = 3)
# 
#   for (i in seq_along(res_idx)) {
#     r <- res_idx[i]
#     txt <- adaptdata$ElementText[r]
#     adaptdata$ResultType_GPT[r] <- classify_result_types_raw(txt)
#     setTxtProgressBar(pb, i)
#   }
#   close(pb)
# } else {
#   cat("No rows with Element == 'Result' found.\n")
# }
# 
# # ---- Quick preview + export (shows EXACT GPT text) ----
# result_preview <- adaptdata %>%
#   filter(grepl("^\\s*result\\s*$", Element, ignore.case = TRUE)) %>%
#   transmute(
#     Country,
#     Document,
#     ElementText = ifelse(
#       is.na(ElementText), "",
#       ifelse(nchar(ElementText) > 120,
#              paste0("<span title='", ElementText, "'>", substr(ElementText, 1, 120), "...</span>"),
#              ElementText)
#     ),
#     ResultType_Manual = ResultType,
#     ResultType_GPT    = ResultType_GPT  # raw GPT output
#   )
# 
# 
# 
# # Save side-by-side comparison for QA
# write.csv(result_preview, "result_type_gpt_preview.csv", row.names = FALSE)
# write.csv(result_preview, "adaptdata_results.csv", row.names = FALSE)
# 

```



```{r,echo=FALSE}
# --- Make "results" = Action Type + (ResultType_GPT) + CI replacements -----------
suppressPackageStartupMessages({
  library(dplyr); library(readr); library(stringr); library(tidyr)
})

# Inputs (your filenames)
path_ci        <- "adaptdata_climateimpact5nano.csv"   # has RowID, ImpactCategories, ClimateImpact_Keywords (authoritative)
path_actionlvl <- "adaptdata_action_level.csv"
path_actiontyp <- "adaptdata_action_type.csv"
path_results   <- "adaptdata_results.csv"              # has ResultType_GPT (and maybe ResultType_Manual)

# Outputs
out_actionlvl  <- "action_level_replaced_impact_ci.csv"
out_actiontyp  <- "action_type_replaced_impact_ci.csv"
out_resultlike <- "action_type_PLUS_results_replaced_impact_ci.csv"  # <-- your “result one”

# ---------- helpers ----------
normalize_element_text <- function(x) {
  x <- as.character(x)
  # pull full text from <span title='...'> if present
  title_val <- stringr::str_match(x, "title=['\"](.*?)['\"]")[,2]
  out <- ifelse(!is.na(title_val), title_val, x)
  out <- gsub("<[^>]+>", " ", out)   # strip tags
  out <- gsub("\\s+", " ", out)      # squeeze whitespace
  trimws(out)
}
drop_cols <- function(df, cols) {
  cols <- intersect(cols, names(df))
  if (length(cols)) dplyr::select(df, -all_of(cols)) else df
}

# ---------- load ----------
ci        <- readr::read_csv(path_ci,        show_col_types = FALSE)
actionlvl <- readr::read_csv(path_actionlvl, show_col_types = FALSE)
actiontyp <- readr::read_csv(path_actiontyp, show_col_types = FALSE)
results   <- readr::read_csv(path_results,   show_col_types = FALSE)

# sanity
need_ci <- c("RowID","ImpactCategories","ClimateImpact_Keywords")
stopifnot(all(need_ci %in% names(ci)))
stopifnot("RowID" %in% names(actionlvl))
stopifnot("RowID" %in% names(actiontyp))
stopifnot(all(c("Country","Document","ElementText") %in% names(results)))

# ---------- authoritative CI map (by RowID) ----------
map_ci <- ci %>%
  mutate(RowID = as.integer(RowID)) %>%
  group_by(RowID) %>%
  summarise(
    ImpactCategories       = dplyr::first(na.omit(ImpactCategories)),
    ClimateImpact_Keywords = dplyr::first(na.omit(ClimateImpact_Keywords)),
    .groups = "drop"
  )

# ---------- 1) Replace/add CI in Action Level & Action Type ----------
actionlvl_out <- actionlvl %>%
  mutate(RowID = as.integer(RowID)) %>%
  drop_cols(c("ImpactCategories","ClimateImpact_Keywords")) %>%
  left_join(map_ci, by = "RowID")

actiontyp_out <- actiontyp %>%
  mutate(RowID = as.integer(RowID)) %>%
  drop_cols(c("ImpactCategories","ClimateImpact_Keywords")) %>%
  left_join(map_ci, by = "RowID")

# ---------- 2) Build “results-like” dataset = Action Type + ResultType_GPT ----------
# Normalize and dedupe results to a key so we can join by text
res_key <- results %>%
  mutate(ElementText_norm = normalize_element_text(ElementText)) %>%
  group_by(Country, Document, ElementText_norm) %>%
  summarise(
    ResultType_GPT    = dplyr::first(na.omit(ResultType_GPT)),
    ResultType_Manual = dplyr::first(na.omit(ResultType_Manual)),
    .groups = "drop"
  )

# Join onto the Action Type file (keep all Action Type rows)
actiontyp_plus_results <- actiontyp_out %>%
  mutate(ElementText_norm = normalize_element_text(ElementText)) %>%
  left_join(res_key, by = c("Country","Document","ElementText_norm")) %>%
  select(-ElementText_norm)


actiontyp_plus_results<-actiontyp_plus_results%>%
  select(-c("...1","...2","...3","...4","ResultType","ResultType_Manual"))

readr::write_csv(actiontyp_plus_results, "adaptdatafinal.csv")





```


```{r,echo=FALSE}
# --- Results: simple bar chart (Input / Process / Output / Outcome) ---

RESULT_LABELS <- c("Input","Process","Output","Outcome")

result_counts <- actiontyp_plus_results %>%
  filter(Element == "Result",
         !is.na(ResultType_GPT),
         nzchar(ResultType_GPT)) %>%
  mutate(ResultType_GPT = str_squish(ResultType_GPT)) %>%
  # map to canonical labels & drop anything off-list (just in case)
  mutate(ResultType_GPT = sapply(ResultType_GPT, function(t) {
    hit <- RESULT_LABELS[tolower(RESULT_LABELS) == tolower(t)]
    ifelse(length(hit), hit, NA_character_)
  })) %>%
  filter(!is.na(ResultType_GPT)) %>%
  mutate(ResultType_GPT = factor(ResultType_GPT, levels = RESULT_LABELS)) %>%
  count(ResultType_GPT, name = "Occurrences") %>%
  arrange(ResultType_GPT) %>%
  mutate(Share = Occurrences / sum(Occurrences),
         Label = paste0(Occurrences, " (", round(Share*100), "%)"))


# horizontal bars with counts + %
p_results <- ggplot(result_counts,
  aes(x = ResultType_GPT, y = Occurrences, fill = ResultType_GPT)) +
  geom_col(width = 0.7, show.legend = FALSE) +
  geom_text(aes(label = Label), hjust = -0.15, size = 4) +
  coord_flip(ylim = c(0, max(result_counts$Occurrences)*1.15)) +
  scale_fill_brewer(palette = "Set2") +  # swap palette if you like
  labs(title = "Result types by count",
       x = NULL, y = "Occurrences") +
  theme_minimal(base_size = 12) +
  theme(axis.text.y = element_text(size = 13),
        axis.text.x = element_text(size = 12))

p_results
```

# link to overview

```{r,echo=FALSE}


adaptdata <- read_csv("adaptdatafinal.csv")

# Ensure both datasets have the Country column
# Check for extra whitespace or formatting issues
adaptdata$Country <- trimws(adaptdata$Country)
overview$Country <- trimws(overview$Country)

# Merge metadata from 'overview' to 'adaptdata' by Country
adaptdata <- merge(adaptdata, overview, by = "Country", all.x = TRUE)

```

# Save data to a local file

```{r,echo=FALSE}
# Convert any list columns to character columns
adaptdata[] <- lapply(adaptdata, function(col) {
  if (is.list(col)) {
    sapply(col, function(x) paste(unlist(x), collapse = "; "))
  } else {
    col
  }
})

write.csv(adaptdata,"AdaptationElementsProcessed.csv")
```


```{r,echo=FALSE}
#------- Missing classifications
# 1) Clean NA-like strings
adapt_clean <- adaptdata %>%
  mutate(across(
    c(SectorType, SectorType_GGA),
    ~na_if(trimws(.), "")
  )) %>%
  mutate(across(
    c(SectorType, SectorType_GGA),
    ~na_if(., "NA")
  ))

# 2) Filter: Element is "System at risk" AND both sector fields are missing
missing_sector_rows <- adapt_clean %>%
  filter(
    Element == "System at risk",
    is.na(SectorType),
    is.na(SectorType_GGA)
  ) %>%
  # pick columns you want in the export (add/remove as needed)
  select(
    Country, Document, Element, ElementLabel,HazardType, SystemType, Sector, SectorType, SectorType_GGA,
    ElementText
  )

write.csv(missing_sector_rows, "missing_ipcc_gga_for_system_at_risk.csv", row.names = FALSE)

# Filter: Element == "System at risk" AND SystemType is missing
system_at_risk_but_no_system <- adapt_clean %>%
  filter(
    Element == "System at risk",
    is.na(SystemType)
  ) %>%
  select(
    Country, Document, Element, ElementText,HazardType,
    SystemType, SectorType, SectorType_GGA
  )

write.csv(system_at_risk_but_no_system,
          "system_at_risk_without_systemtype.csv",
          row.names = FALSE)

```

# System at risk → IPCC sector → GGA

```{r,echo=FALSE}
# -------- Build triples (System -> IPCC -> GGA) --------

library(dplyr)
library(tidyr)
library(stringr)

# Clean NA-like strings and whitespace
adaptdata <- adaptdata %>%
  mutate(across(
    c(SystemType, SectorType, SectorType_GGA),
    ~na_if(trimws(.), "")
  )) %>%
  mutate(across(
    c(SystemType, SectorType, SectorType_GGA),
    ~na_if(., "NA")  # Handle "NA" strings
  ))

triples <- adaptdata %>%
  filter(
    !is.na(SystemType),
    !is.na(SectorType),
    !is.na(SectorType_GGA)
  ) %>%
  count(SystemType, SectorType, SectorType_GGA, name = "value") %>%
  arrange(desc(value))

edges_sys_sector <- triples %>%
  count(from = SystemType, to = SectorType, wt = value, name = "value")
edges_sector_gga <- triples %>%
  count(from = SectorType, to = SectorType_GGA, wt = value, name = "value")


```

# Remarks

- **Hazard Type:**  
  A few additional climate-related keywords were added to improve coverage. For example, terms like *“dry days”*, *“surface air temperature”*, and *“heat stress”* appeared frequently and were added under appropriate hazard categories.  
  However, we also encountered terms such as *“runoff”*, *“solar radiation”*, and *“winds”* that were not clearly assignable to a single hazard category and remain unclassified for now.

- **System at Risk:**  
  We expanded the keyword list to better capture common themes in the data:
  
  - **Coastal:** Added terms like *“eutrophication”* and *“algal bloom”*, which frequently occur in the context of coastal or marine environmental issues.
  - **Livelihoods and Poverty:** Included tourism-related terms such as *“tourism”*, *“tourists”*, *“ski”*, *“vacation”*, and *“outdoor activities”*, to capture socio-economic impacts on income and livelihoods.
  - **Infrastructure and Services:** Extended to include infrastructure and utility-related keywords such as *“power”*, *“hydropower”*, *“railways”*, *“port”*, *“industry”*, *“industries”*, and *“material”*.
  - **Health:** Added terms like *“pollution”*, *“deaths”*, *“human lives”*, and *“life expectancy”* to better reflect public health consequences mentioned in the data.
 

  ⚠️ **Note:** We observed several rows tagged as *System at risk* that likely describe hazards rather than systems, as they do not mention any system but describe a climate hazard (e.g., drought or storms). This may warrant a second look or reclassification.
  Some examples 
  
  - Climate change is projected to cause longer and dryer dry seasons and shorter rainy seasons
  - Increased frequency and intensity of flooding in vulnerable areas
  - Another precursor to higher drought incidence will be the expected change in rainfall variability, with the number of rainfall days decreasing, especially in spring and summer, while the intensity of individual rainfall events increases.
  - Seychelles is projected to experience increased average temperatures and changes in rainfall patterns, leading to more frequent and intense rainfall events, including flooding and landslides, and potentially longer periods of drought


- **Sector Tagging (IPCC & GGA):**  
  Sector tagging shows **strong overall coverage**, particularly due to the mapping from both direct sector mentions and `SystemType`.  
  That said, a number of rows still remain **unclassified**, either because sector information wasn't mentioned explicitly in the original source or because it was too ambiguous to map reliably. It's unclear whether these untagged rows should be a concern, but they could be flagged for manual review depending on use case.
